---
title: "Practicum_Final_Analysis"
output: html_document
date: "2026-04-19"
---

```{r setup, include=FALSE}
library(survminer)
library(cobalt)
library(survival)
library(SimDesign)
library(writexl)
library(readxl)
library(tidyr)
library(stringr)
library(gt)
setwd("~/Documents/School/Practicum")
```
 
############
Poster Data
 
```{r}
results_poster_full <- read.csv("outputs/poster_data_long.csv")
```
 
Change the dataset to long format for easier formatting. 
```{r}
poster_long <- results_poster_full %>%
  pivot_longer(
    cols = matches("^(oracle|mcar|mar|mnar)\\.(mean\\.hr|median\\.hr|bias|rmse|coverage|power)\\.(dfs|os)$"),
    names_to = c("missing_mechanism", "metric", "outcome"),
    names_pattern = "^(oracle|mcar|mar|mnar)\\.(mean\\.hr|median\\.hr|bias|rmse|coverage|power)\\.(dfs|os)$",
    values_to = "value"
  ) %>%
  mutate(
    missing_mechanism = toupper(missing_mechanism),
    outcome = toupper(outcome)
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  )

poster_long <- poster_long %>%
  mutate(
    missing_mechanism = factor(missing_mechanism, levels = c("ORACLE", "MCAR", "MAR", "MNAR")),
    outcome = factor(outcome, levels = c("DFS", "OS")),
    causal_method = factor(causal_method, levels = c("match", "ebal", "iptw")),
    input_ss = factor(input_ss),
    input_mis_subj_lab = paste0(100 * input_mis_subj, "%"),
    missing_method = case_when(
      missing_method == "complete_case" ~ "Complete Case",
      missing_method == "multiple_imputation" ~ "Multiple Imputation"
    )
  )
```

# Main table

```{r}
table_main <- poster_long %>%
  dplyr::select(
    outcome, missing_mechanism,
    input_ss, input_hr_dfs, input_hr_os, input_mis_subj,
    causal_method, missing_method,
    mean.hr, median.hr, bias, rmse, coverage
  ) %>%
  arrange(outcome, missing_mechanism, input_mis_subj, causal_method, input_ss)

# Returns the average of the simulation data based on outcome, missing mechanism, missing %, causal method and missing method
# 
table_avg <- poster_long %>%
  group_by(outcome, missing_mechanism, input_mis_subj, causal_method, missing_method) %>%
  summarise(
    mean_hr = mean(mean.hr, na.rm = TRUE),
    median_hr = mean(median.hr, na.rm = TRUE),
    mean_bias = mean(bias, na.rm = TRUE),
    mean_rmse = mean(rmse, na.rm = TRUE),
    mean_coverage = mean(coverage, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(outcome, missing_mechanism, input_mis_subj, causal_method)
```

#Plots

```{r}
ggplot(
  table_avg %>% filter(missing_mechanism != "ORACLE"),
  aes(x = input_mis_subj, y = mean_bias, color = causal_method, group = causal_method)
) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_grid(missing_method ~ missing_mechanism) +
  labs(
    x = "Missingness Proportion",
    y = "Average Bias",
    subtitle = "Dashed line = zero bias",
    color = "Causal Method",
    title = "Bias Across Missingness Mechanisms"
  ) +
  theme_sim()

plot_base %>%
  filter(outcome == "DFS") %>%
  ggplot(aes(x = input_mis_subj, y = mean_bias,
             color = causal_method, group = causal_method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_grid(missing_method ~ missing_mechanism) +
  scale_color_manual(values = method_colors, name = "Causal method") +
  scale_x_continuous(breaks = c(0, 0.25, 0.50),
                     labels = c("0%", "25%", "50%")) +
  labs(title    = "Bias across missingness mechanisms",
       subtitle = "Dashed line = zero bias",
       x        = "Missingness proportion",
       y        = "Mean bias (log HR)") +
  theme_sim()

```
```{r}
ggplot(
  table_avg %>% filter(missing_mechanism != "ORACLE"),
  aes(x = input_mis_subj, y = mean_rmse, color = causal_method, group = causal_method)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_grid(outcome ~ missing_mechanism) +
  labs(
    x = "Missingness proportion",
    y = "Average RMSE",
    color = "Causal method",
    title = "RMSE across missingness mechanisms"
  ) +
  theme_bw()
```

```{r}
ggplot(
  table_avg %>% filter(missing_mechanism != "ORACLE", outcome == "DFS"),
  aes(x = input_mis_subj, y = mean_coverage, color = causal_method, group = causal_method)
) +
  geom_hline(yintercept = 0.95, linetype = 2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_grid(missing_method ~ missing_mechanism) +
  labs(
    x = "DFS Missingness proportion",
    y = "Average coverage",
    color = "Causal method",
    title = "Coverage across missingness mechanisms"
  ) +
  theme_bw()

ggplot(
  table_avg %>% filter(missing_mechanism != "ORACLE", outcome == "OS"),
  aes(x = input_mis_subj, y = mean_coverage, color = causal_method, group = causal_method)
) +
  geom_hline(yintercept = 0.95, linetype = 2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_grid(missing_method ~ missing_mechanism) +
  labs(
    x = "OS Missingness proportion",
    y = "Average coverage",
    color = "Causal method",
    title = "Coverage across missingness mechanisms"
  ) +
  theme_bw()
```



The most interesting findings are:
1. Ebal is the most robust method — consistently near 0.95 across all mechanisms and missingness levels under complete case
2. Complete case handles MCAR and MAR reasonably well — coverage stays near 0.95, suggesting complete case analysis is not dramatically worse than MI for these mechanisms
3. MNAR degrades complete case performance — the expected finding, validates the simulation is working
4. MI with matching is problematic — overcoverage suggests the combination of full matching weights and Rubin's rules inflates uncertainty
5. The increasing coverage with missingness under MI is counterintuitive and worth investigating — it may be an artifact of the MI CIs being wider (more conservative) as missingness increases, which mechanically inflates coverage even if point estimates degrade

```{r}
tbl_data <- data.frame(
  Factor = c(
    "Sample size",
    "True HR (DFS/OS)",
    "Missingness proportion",
    "Missingness mechanism",
    "Causal method",
    "Missing data strategy"
  ),
  Levels = c(
    "500, 1000, 2000",
    "1.0, 1.5, 2.0",
    "0%, 25%, 50%",
    "MCAR, MAR, MNAR",
    "IPTW, ebal, matching",
    "Complete case, MI"
  )
)

gt(tbl_data) %>%
  tab_header(
    title = "Sample Size / Conditions Table"
  )
```


 
 