---
title: "Practicum_Causal_Practise"
output: pdf_document
date: "2026-02-24"
---

```{r setup, include=FALSE}
library(dplyr)
library(WeightIt)
library(MatchIt)
library(survminer)
library(cobalt)
library(survival)
library(SimDesign)
library(writexl)
setwd("~/Documents/School/Practicum")
```
# Prep Data
```{r}
pred.vars <- c("Age_Dx.dich", "pT_7th", "pN_7th", "ENE_", "PORT", "Max_tumor_size","Total_LN_positive_")
cat.vars <- c("Age_Dx.dich", "pT_7th", "pN_7th", "ENE_", "PORT")
cont.vars <- c("Max_tumor_size","Total_LN_positive_")

# Function to prepare data
prep_data <- function(data) {
  cat("Starting data preparation...\n")
  
  # Check input data immediately
  if(nrow(data) == 0) {
    stop("PREP_DATA ERROR: Input data is empty")
  }
  
  # Remove rows with missing Treatment/outcome data - FAIL if critical variables missing
  if(!all(c("DFStime", "DFS") %in% colnames(data))) {
    stop("PREP_DATA ERROR: Missing required survival variables DFStime or DFS")
  }
  
  data <- data[!is.na(data$DFStime) & !is.na(data$DFS), ]
  
  if(nrow(data) == 0) {
    stop("PREP_DATA ERROR: No valid survival data remaining after removing NAs")
  }
  
  # CRITICAL FIX: Create Treatment variable from engraftment1
  if(!"engraftment1" %in% colnames(data)) {
    stop("PREP_DATA ERROR: engraftment1 variable not found in data - this is the source treatment variable")
  }
  
  # Check what values exist in engraftment1
  unique_engraft <- unique(data$engraftment1[!is.na(data$engraftment1)])
  cat("Found engraftment1 values:", paste(unique_engraft, collapse = ", "), "\n")
  
  # Create Treatment variable (1 = Engrafter, 0 = Non-Engrafter)
  # Based on project knowledge: "Engrafter" = 1, "Non-Engrafter" = 0
  data$Treatment <- ifelse(data$engraftment1 == "Engrafter", 1, 0)
  
  # Validate Treatment creation
  if(all(is.na(data$Treatment))) {
    stop("PREP_DATA ERROR: Treatment variable creation failed - all values are NA")
  }
  
  treatment_counts <- table(data$Treatment, useNA = "always")
  cat("Treatment assignment counts:\n")
  print(treatment_counts)
  
  if(sum(!is.na(data$Treatment)) < 20) {
    stop("PREP_DATA ERROR: Insufficient valid Treatment assignments")
  }
  
  # Create binary versions - FAIL immediately if source variables don't exist
  if(!"Age_Dx" %in% colnames(data)) {
    stop("PREP_DATA ERROR: Age_Dx variable not found in data")
  }
  
  # Convert Age_Dx to numeric - fail if conversion impossible
  data$Age_Dx <- as.numeric(as.character(data$Age_Dx))
  if(all(is.na(data$Age_Dx))) {
    stop("PREP_DATA ERROR: Age_Dx could not be converted to numeric")
  }
  
  data$Age_Dx_bin <- ifelse(data$Age_Dx >= median(data$Age_Dx, na.rm = TRUE), 1, 0)
  
  # Process other variables - FAIL if expected variables missing
  var_mappings <- list(
    pT_7th_bin = "pT_7th",
    pN_7th_bin = "pN_7th", 
    ENE_bin = "ENE_",
    PORT_bin = "PORT"
  )
  
  for(new_var in names(var_mappings)) {
    old_var <- var_mappings[[new_var]]
    
    if(!old_var %in% colnames(data)) {
      stop(paste("PREP_DATA ERROR: Required variable", old_var, "not found in data"))
    }
    
    if(is.character(data[[old_var]]) || is.factor(data[[old_var]])) {
      # Convert categorical to binary
      unique_vals <- unique(data[[old_var]][!is.na(data[[old_var]])])
      if(length(unique_vals) < 2) {
        stop(paste("PREP_DATA ERROR: Variable", old_var, "has insufficient unique values:", paste(unique_vals, collapse = ", ")))
      }
      
      # For specific OSCC variables, use proper coding:
      if(old_var == "pT_7th") {
        # T3-4 = 1, T1-2 = 0 (based on project knowledge)
        data[[new_var]] <- ifelse(data[[old_var]] %in% c("T3-4"), 1, 0)
      } else if(old_var == "pN_7th") {
        # N2-3 = 1, N0-1 = 0 (based on project knowledge)
        data[[new_var]] <- ifelse(data[[old_var]] %in% c("N2-3"), 1, 0)
      } else if(old_var == "ENE_") {
        # Yes = 1, No = 0
        data[[new_var]] <- ifelse(data[[old_var]] == "Yes", 1, 0)
      } else if(old_var == "PORT") {
        # Yes = 1, No = 0
        data[[new_var]] <- ifelse(data[[old_var]] == "Yes", 1, 0)
      } else {
        # Default: first unique value = 0, others = 1
        data[[new_var]] <- ifelse(data[[old_var]] == unique_vals[1], 0, 1)
      }
    } else {
      # Convert numeric to binary
      numeric_vals <- as.numeric(as.character(data[[old_var]]))
      if(all(is.na(numeric_vals))) {
        stop(paste("PREP_DATA ERROR: Variable", old_var, "could not be converted to numeric"))
      }
      data[[new_var]] <- ifelse(numeric_vals > 0, 1, 0)
    }
    cat("Created", new_var, "from", old_var, "\n")
  }
  
  # Ensure continuous variables exist and are numeric - FAIL if missing
  for(var in c("Max_tumor_size", "Total_LN_positive_")) {
    if(!var %in% colnames(data)) {
      stop(paste("PREP_DATA ERROR: Required continuous variable", var, "not found"))
    }
    
    data[[var]] <- as.numeric(as.character(data[[var]]))
    if(all(is.na(data[[var]]))) {
      stop(paste("PREP_DATA ERROR: Variable", var, "could not be converted to numeric"))
    }
    
    # Replace negative values with 0 (this is acceptable data cleaning)
    data[[var]][data[[var]] < 0 | is.na(data[[var]])] <- 0
  }
  
  # Ensure survival variables are properly formatted - FAIL if invalid
  required_vars <- c("DFStime", "DFS", "OStime", "OS")
  missing_vars <- required_vars[!required_vars %in% colnames(data)]
  if(length(missing_vars) > 0) {
    stop(paste("PREP_DATA ERROR: Missing required survival variables:", paste(missing_vars, collapse = ", ")))
  }
  
  # Convert survival variables to numeric
  for(var in required_vars) {
    data[[var]] <- as.numeric(as.character(data[[var]]))
    if(all(is.na(data[[var]]))) {
      stop(paste("PREP_DATA ERROR: Survival variable", var, "could not be converted to numeric"))
    }
  }
  
  # Remove invalid survival data
  before_count <- nrow(data)
  data <- data[data$DFStime > 0 & data$OStime > 0 & 
               !is.na(data$DFStime) & !is.na(data$OStime), ]
  after_count <- nrow(data)
  
  if(after_count == 0) {
    stop("PREP_DATA ERROR: No valid survival times remaining (all <= 0 or NA)")
  }
  
  if(after_count < before_count * 0.5) {
    stop(paste("PREP_DATA ERROR: Lost too much data in survival cleaning:", before_count, "->", after_count, "rows"))
  }
  
  # Final validation of Treatment variable
  final_treatment_table <- table(data$Treatment, useNA = "always")
  cat("Final Treatment distribution:\n")
  print(final_treatment_table)
  
  if(sum(data$Treatment == 1, na.rm = TRUE) < 5) {
    stop("PREP_DATA ERROR: Insufficient treated subjects (engrafters)")
  }
  
  if(sum(data$Treatment == 0, na.rm = TRUE) < 5) {
    stop("PREP_DATA ERROR: Insufficient control subjects (non-engrafters)")
  }
  
  cat("Data preparation completed successfully:", after_count, "valid rows\n")
  cat("Treatment prevalence:", round(mean(data$Treatment, na.rm = TRUE), 3), "\n")
  return(data)
}

original_data <- read.csv("predata_s2.csv")

data <- prep_data(original_data)

pred.vars.processed <- c("Age_Dx_bin", "pT_7th_bin", "pN_7th_bin", "ENE_bin", "PORT_bin", "Max_tumor_size", "Total_LN_positive_")
cat.vars.processed <- c("Age_Dx_bin", "pT_7th_bin", "pN_7th_bin", "ENE_bin", "PORT_bin")
cont.vars.processed <- c("Max_tumor_size", "Total_LN_positive_")

# Create reference data for simulation - FAIL if insufficient data
ref.data.processed <- data[, pred.vars.processed, drop = FALSE]
ref.data.processed <- ref.data.processed[complete.cases(ref.data.processed), , drop = FALSE]
```

We will simulate the variables in a similar but distinctly different manner defined by `ff.dummy.cdf`. This uses the CDF of a uniform instead to perform the underlying structure of the data since a MVN assumption may not hold. 

```{r empirical_distribution_functions}
# Empirical distribution functions - NO FALLBACKS, direct errors
# Compute mu and sigma for variables
ff.mean.sigma <- function(x) {
  if(!is.data.frame(x)) {
    x <- as.data.frame(x)
  }
  
  if(nrow(x) == 0 || ncol(x) == 0) {
    stop("FF.MEAN.SIGMA ERROR: Input data frame has no rows or columns")
  }
  
  if(is.null(colnames(x)) || any(colnames(x) == "")) {
    stop("FF.MEAN.SIGMA ERROR: Missing or empty column names detected")
  }
  
  cat("Computing empirical distributions for variables:", paste(colnames(x), collapse = ", "), "\n")
  
  # Ensure all data is numeric - FAIL if conversion impossible
  for(i in 1:ncol(x)) {
    original_values <- x[, i]
    x[, i] <- as.numeric(as.character(x[, i]))
    
    if(all(is.na(x[, i]))) {
      stop(paste("FF.MEAN.SIGMA ERROR: Column", colnames(x)[i], "could not be converted to numeric"))
    }
    
    # Check for non-finite values
    non_finite_count <- sum(!is.finite(x[, i]))
    if(non_finite_count > nrow(x) * 0.5) {
      stop(paste("FF.MEAN.SIGMA ERROR: Column", colnames(x)[i], "has too many non-finite values:", non_finite_count, "/", nrow(x)))
    }
    
    # Replace non-finite with 0 (acceptable for binary variables)
    x[!is.finite(x[, i]), i] <- 0
  }
  
  # Calculate variance and means
  mu <- colMeans(x, na.rm = TRUE)
  sigma <- var(x, use = "pairwise.complete.obs")
  
  if(any(is.na(sigma)) || any(is.na(mu))) {
    stop("FF.MEAN.SIGMA ERROR: NA values detected in computed variance or means")
  }
  
  # Check for positive definite covariance matrix - FAIL if not fixable
  # Positive eigvenvalues => positive definite covariance
  eigenvals <- eigen(sigma, only.values = TRUE)$values
  min_eigenval <- min(eigenvals)
  
  if(min_eigenval <= 1e-8) {
    cat("Warning: Non-positive definite covariance matrix detected. Min eigenvalue:", min_eigenval, "\n")
    # Add minimal regularization
    regularization <- max(1e-6, abs(min_eigenval) + 1e-6)
    sigma <- sigma + diag(regularization, nrow(sigma))
    cat("Added regularization:", regularization, "\n")
    
    # Check if regularization worked
    new_eigenvals <- eigen(sigma, only.values = TRUE)$values
    if(min(new_eigenvals) <= 0) {
      stop("FF.MEAN.SIGMA ERROR: Could not create positive definite covariance matrix even after regularization")
    }
  }
  
  return(list(mu = mu, sigma = sigma))
}

# Binary variable conversion - NO FALLBACKS
# Convert a simulated column into a binary column
ff.dummy <- function(sim_data, empirical_dist, var_name) {
  if(!var_name %in% colnames(sim_data)) {
    stop(paste("FF.DUMMY ERROR: Variable", var_name, "not found in simulation data"))
  }
  
  if(!var_name %in% names(empirical_dist$mu)) {
    stop(paste("FF.DUMMY ERROR: Variable", var_name, "not found in empirical distribution"))
  }
  
  empirical_prop <- empirical_dist$mu[var_name]
  
  if(is.na(empirical_prop) || !is.finite(empirical_prop)) {
    stop(paste("FF.DUMMY ERROR: Invalid empirical proportion for", var_name, ":", empirical_prop))
  }
  
  if(empirical_prop < 0 || empirical_prop > 1) {
    stop(paste("FF.DUMMY ERROR: Empirical proportion out of bounds [0,1] for", var_name, ":", empirical_prop))
  }
  
  simulated_vals <- as.numeric(sim_data[, var_name])
  
  if(all(is.na(simulated_vals))) {
    stop(paste("FF.DUMMY ERROR: All simulated values for", var_name, "are NA"))
  }
  
  # Calculate threshold based on empirical proportion
  threshold <- quantile(simulated_vals, 1 - empirical_prop, na.rm = TRUE)
  
  sim_data[, var_name] <- ifelse(simulated_vals > threshold, 1, 0)
  
  simulated_prop <- mean(sim_data[, var_name], na.rm = TRUE)
  cat("Converted", var_name, "to binary: empirical prop =", round(empirical_prop, 3), 
      ", simulated prop =", round(simulated_prop, 3), "\n")
  
  return(sim_data)
}

# Instead convert the normals into uniform through the normal CDF:
ff.dummy_cdf <- function(sim_data, empirical_dist, var_name) {
  if(!var_name %in% colnames(sim_data)) {
    stop(paste("FF.DUMMY_CDF ERROR: Variable", var_name, "not found in simulation data"))
  }
  
  if(!var_name %in% names(empirical_dist$mu)) {
    stop(paste("FF.DUMMY_CDF ERROR: Variable", var_name, "not found in empirical distribution"))
  }
  
  if(!(var_name %in% rownames(empirical_dist$sigma))) {
    stop(paste("FF.DUMMY_CDF ERROR: Variable", var_name, "not found in covariance matrix"))
  }
  
  empirical_prop <- empirical_dist$mu[var_name]
  
  if(is.na(empirical_prop) || !is.finite(empirical_prop) || empirical_prop < 0 || empirical_prop > 1) {
    stop(paste("FF.DUMMY_CDF ERROR: Invalid empirical proportion for", var_name, ":", empirical_prop))
  }
  
  z <- as.numeric(sim_data[[var_name]])
  
  if(all(is.na(z))) {
    stop(paste("FF.DUMMY_CDF ERROR: All simulated values for", var_name, "are NA"))
  }
  
  mu_z <- empirical_dist$mu[var_name]
  sd_z <- sqrt(empirical_dist$sigma[var_name, var_name])
  
  if(!is.finite(sd_z) || sd_z <= 0) {
    stop(paste("FF.DUMMY_CDF ERROR: Invalid latent SD for", var_name, ":", sd_z))
  }
  
  # Convert latent normal values to U(0,1) using the normal CDF
  u <- pnorm(z, mean = mu_z, sd = sd_z)
  
  # Assign binary values using the empirical prevalence
  sim_data[[var_name]] <- ifelse(u <= empirical_prop, 1, 0)
  
  simulated_prop <- mean(sim_data[[var_name]], na.rm = TRUE)
  cat("Converted", var_name,
      "to binary via CDF: empirical prop =", round(empirical_prop, 3),
      ", simulated prop =", round(simulated_prop, 3), "\n")
  
  return(sim_data)
}

# Simulate covariates based on the empirical distribution 
# Covariate simulation - NO FALLBACKS
simulate_covariates_empirical <- function(empirical_dist, n_sim, variables, cat_vars, cont_vars) {
  cat("\n=== SIMULATING COVARIATES USING EMPIRICAL DISTRIBUTIONS ===\n")
  cat("Sample size:", n_sim, "\n")
  cat("Variables:", paste(variables, collapse = ", "), "\n")
  
  if(n_sim <= 0) {
    stop("SIMULATE_COVARIATES ERROR: Invalid sample size:", n_sim)
  }
  
  if(length(variables) == 0) {
    stop("SIMULATE_COVARIATES ERROR: No variables specified for simulation")
  }
  
  # Check empirical distribution validity
  missing_vars <- variables[!variables %in% names(empirical_dist$mu)]
  if(length(missing_vars) > 0) {
    stop(paste("SIMULATE_COVARIATES ERROR: Variables missing from empirical distribution:", paste(missing_vars, collapse = ", ")))
  }
  
  if(any(!is.finite(empirical_dist$mu[variables]))) {
    stop("SIMULATE_COVARIATES ERROR: Non-finite means detected in empirical distribution")
  }
  
  cat("Using empirical means:", paste(round(empirical_dist$mu[variables], 3), collapse = ", "), "\n")
  
  # Extract covariance matrix for specified variables
  cov_matrix <- empirical_dist$sigma[variables, variables]
  
  # Ensure matrix is symmetric
  if(!isSymmetric(cov_matrix)) {
    cov_matrix <- (cov_matrix + t(cov_matrix)) / 2
  }
  
  # Generate multivariate normal data - FAIL if impossible
  sim_data <- as.data.frame(MASS::mvrnorm(n = n_sim, 
                                         mu = empirical_dist$mu[variables], 
                                         Sigma = cov_matrix))
  colnames(sim_data) <- variables
  cat("Generated", nrow(sim_data), "subjects with", ncol(sim_data), "variables\n")
  
  # Transform binary variables
  for(var in cat_vars) {
    if(var %in% variables) {
      sim_data <- ff.dummy_cdf(sim_data, empirical_dist, var)
    }
  }
  cat("Sucessfully transformed category variables: ", paste(cat_vars, collapse = ", "), "\n")
  
  # Ensure continuous variables are non-negative
  for(var in cont_vars) {
    if(var %in% variables) {
      negative_count <- sum(sim_data[, var] < 0)
      if(negative_count > 0) {
        cat("Applied non-negativity constraint to", var, "(", negative_count, "negative values)\n")
      }
      sim_data[, var] <- pmax(0, as.numeric(sim_data[, var]))
    }
  }
  
  cat("=== COVARIATE SIMULATION COMPLETED ===\n\n")
  return(sim_data)
}
```


```{r treatment_estimation}
# Treatment model estimation
# Fits two logistic regressions for treatment and PORT
estimate_treatment_models <- function(data) {
  cat("Extracting empirical treatment assignment patterns from OSCC data...\n")
  
  if(nrow(data) == 0) {
    stop("TREATMENT_MODELS ERROR: Input data is empty")
  }
  
  # Check if required variables exist
  required_vars <- c("Age_bin_DX", "ENE_bin", "pN_7th_bin", "pT_7th_bin", "Max_tumor_size", "Total_LN_positive_")
  available_vars <- intersect(required_vars, colnames(data))
  
  if(length(available_vars) == 0) {
    stop("TREATMENT_MODELS ERROR: No predictor variables available for treatment models")
  }
  
  cat("Using variables for treatment model:", paste(available_vars, collapse = ", "), "\n")
  
  # Model 1: Treatment patterns (using "Treatment" variable if available)
  engraftment_model <- NULL
  if("Treatment" %in% colnames(data)) {
    formula_str <- paste("Treatment ~", paste(available_vars, collapse = " + "))
    treatment_formula <- as.formula(formula_str)
    
    cat("Fitting engraftment model...\n")
    engraftment_model <- glm(treatment_formula, data = data, family = binomial())
    
    if(is.null(engraftment_model) || any(is.na(coef(engraftment_model)))) {
      stop("TREATMENT_MODELS ERROR: Engraftment model failed or produced NA coefficients")
    }
    
    cat("Empirical treatment model:\n")
    print(round(coef(engraftment_model), 3))
    cat("Observed treatment rate:", round(mean(data$Treatment, na.rm = TRUE), 3), "\n\n")
  } else {
    stop("TREATMENT_MODELS ERROR: Treatment variable not found in data")
  }
  
  # Model 2: PORT assignment patterns
  PORT_model <- NULL
  if("PORT_bin" %in% colnames(data)) {
    formula_str <- paste("PORT_bin ~", paste(available_vars, collapse = " + "))
    PORT_formula <- as.formula(formula_str)
    
    cat("Fitting PORT model...\n")
    PORT_model <- glm(PORT_formula, data = data, family = binomial())
    
    if(is.null(PORT_model) || any(is.na(coef(PORT_model)))) {
      stop("TREATMENT_MODELS ERROR: PORT model failed or produced NA coefficients")
    }
    
    cat("Empirical PORT model:\n")
    print(round(coef(PORT_model), 3))
    cat("Observed PORT rate:", round(mean(data$PORT_bin, na.rm = TRUE), 3), "\n\n")
  } else {
    stop("TREATMENT_MODELS ERROR: PORT_bin variable not found in data")
  }
  
  return(list(
    engraftment_model = coef(engraftment_model),
    PORT_model = coef(PORT_model)
  ))
}
```

```{r treatment_assignment}
simulate_treatment_assignment <- function(sim_data, treatment_type, treatment_models) {
  
  if(nrow(sim_data) == 0) {
    stop("TREATMENT_ASSIGNMENT ERROR: Simulation data is empty")
  }
  
  if(treatment_type == "engraftment") {
    model_coefs <- treatment_models$engraftment_model  
  } else if(treatment_type == "PORT") {
    model_coefs <- treatment_models$PORT_model  
  } else {
    stop(paste("TREATMENT_ASSIGNMENT ERROR: Unknown treatment type:", treatment_type))
  }
  
  if(is.null(model_coefs) || length(model_coefs) == 0) {
    stop(paste("TREATMENT_ASSIGNMENT ERROR: No coefficients available for treatment type:", treatment_type))
  }
  
  cat("Using EMPIRICAL treatment coefficients from OSCC data\n")
  cat("Coefficients: Intercept =", round(model_coefs[1], 3))
  if(length(model_coefs) > 1) {
    cat(", Others =", paste(round(model_coefs[2:min(3, length(model_coefs))], 3), collapse = ", "))
  }
  cat("\n")
  
  # Find available variables for treatment model
  available_vars <- intersect(c("ENE_bin", "pN_7th_bin", "pT_7th_bin", "Max_tumor_size", "Total_LN_positive_"), 
                              colnames(sim_data))
  
  if(length(available_vars) == 0) {
    stop("TREATMENT_ASSIGNMENT ERROR: No predictor variables available in simulation data")
  }
  
  # Create formula dynamically
  formula_str <- paste("~", paste(available_vars, collapse = " + "))
  
  # Scale continuous variables if they exist
  sim_data_scaled <- sim_data
  for(var in available_vars) {
    if(var %in% c("Max_tumor_size", "Total_LN_positive_")) {
      scaled_vals <- scale(sim_data[[var]])[,1]
      if(any(!is.finite(scaled_vals))) {
        stop(paste("TREATMENT_ASSIGNMENT ERROR: Scaling produced non-finite values for", var))
      }
      sim_data_scaled[[var]] <- scaled_vals
    }
  }
  
  # Create design matrix
  X_treatment <- model.matrix(as.formula(formula_str), data = sim_data_scaled)
  
  if(ncol(X_treatment) != length(model_coefs)) {
    stop(paste("TREATMENT_ASSIGNMENT ERROR: Design matrix columns (", ncol(X_treatment), 
               ") don't match coefficient length (", length(model_coefs), ")"))
  }
  
  # Calculate linear predictor
  linear_pred <- X_treatment %*% model_coefs
  
  if(any(!is.finite(linear_pred))) {
    stop("TREATMENT_ASSIGNMENT ERROR: Linear predictor contains non-finite values")
  }
  
  treatment_prob <- plogis(linear_pred)
  treatment <- rbinom(nrow(sim_data), 1, treatment_prob)
  
  cat("Treatment assignment summary:\n")
  cat("  Treatment prevalence:", round(mean(treatment), 3), "\n")
  cat("  Propensity score range:", round(min(treatment_prob), 3), "-", round(max(treatment_prob), 3), "\n")
  
  return(treatment)
}
```

```{r baseline_models}
# Baseline model estimation
# Fit baseline models for DFS and OS (Cox models and Weibull models)
estimate_baseline_models <- function(data) {
  
  if(nrow(data) == 0)
    stop("BASELINE_MODELS ERROR: Input data is empty")
  
  required_vars <- c("DFStime", "DFS", "OStime", "OS")
  missing_vars  <- required_vars[!required_vars %in% colnames(data)]
  if(length(missing_vars) > 0)
    stop(paste("BASELINE_MODELS ERROR: Missing required survival variables:", 
               paste(missing_vars, collapse = ", ")))
  
  available_vars <- intersect(c("Age_Dx_bin", "pT_7th_bin", "pN_7th_bin",
                                 "ENE_bin", "PORT_bin", "Max_tumor_size", 
                                 "Total_LN_positive_"),
                               colnames(data))
  
  if(length(available_vars) == 0)
    stop("BASELINE_MODELS ERROR: No predictor variables available")
  
  cat("Using predictor variables:", paste(available_vars, collapse = ", "), "\n")
  
  # Shared formulas — reused for both Cox and Weibull
  dfs_formula <- as.formula(paste("Surv(DFStime, DFS) ~", 
                                   paste(available_vars, collapse = " + ")))
  os_formula  <- as.formula(paste("Surv(OStime, OS) ~",  
                                   paste(available_vars, collapse = " + ")))
  
  # Cox models 
  cat("Fitting DFS Cox model...\n")
  dfs_model <- coxph(dfs_formula, data = data)
  if(is.null(dfs_model) || any(is.na(coef(dfs_model))))
    stop("BASELINE_MODELS ERROR: DFS Cox model failed or produced NA coefficients")
  
  cat("Fitting OS Cox model...\n")
  os_model <- coxph(os_formula, data = data)
  if(is.null(os_model) || any(is.na(coef(os_model))))
    stop("BASELINE_MODELS ERROR: OS Cox model failed or produced NA coefficients")
  
  # Weibull models — with covariates for correct baseline
  # Using same formula as Cox ensures lambda/gamma correspond to 
  # the baseline hazard at the reference covariate profile
  cat("Fitting DFS Weibull model...\n")
  dfs_survreg <- survreg(dfs_formula, data = data, dist = "weibull")
  if(is.null(dfs_survreg) || is.na(dfs_survreg$coefficients["(Intercept)"]) || 
     is.na(dfs_survreg$scale))
    stop("BASELINE_MODELS ERROR: DFS Weibull model failed or produced NA parameters")
  
  dfs_lambda <- exp(-dfs_survreg$coefficients["(Intercept)"] / dfs_survreg$scale)
  dfs_gamma <- 1 / dfs_survreg$scale
  
  if(!is.finite(dfs_lambda) || !is.finite(dfs_gamma) || 
     dfs_lambda <= 0 || dfs_gamma <= 0)
    stop(paste("BASELINE_MODELS ERROR: Invalid DFS Weibull parameters: lambda =", 
               dfs_lambda, ", gamma =", dfs_gamma))
  
  cat("Fitting OS Weibull model...\n")
  os_survreg <- survreg(os_formula, data = data, dist = "weibull")
  if(is.null(os_survreg) || is.na(os_survreg$coefficients["(Intercept)"]) || 
     is.na(os_survreg$scale))
    stop("BASELINE_MODELS ERROR: OS Weibull model failed or produced NA parameters")
  
  os_lambda <- exp(-os_survreg$coefficients["(Intercept)"] / os_survreg$scale)
  os_gamma <- 1 / os_survreg$scale
  
  if(!is.finite(os_lambda) || !is.finite(os_gamma) || 
     os_lambda <= 0 || os_gamma <= 0)
    stop(paste("BASELINE_MODELS ERROR: Invalid OS Weibull parameters: lambda =", 
               os_lambda, ", gamma =", os_gamma))
  
  cat("Baseline models fitted successfully\n")
  cat("DFS — lambda:", round(dfs_lambda, 4), "gamma:", round(dfs_gamma, 4), "\n")
  cat("OS  — lambda:", round(os_lambda,  4), "gamma:", round(os_gamma,  4), "\n")
  
  return(list(
    dfs_cox = dfs_model,
    os_cox = os_model,
    dfs_lambda = dfs_lambda,
    dfs_gamma = dfs_gamma,
    os_lambda = os_lambda,
    os_gamma = os_gamma,
    dfs_coef = coef(dfs_model),
    os_coef = coef(os_model)
  ))
}
```

```{r simulated_outcomes}
simulate_outcomes <- function(dataset, treatment,
                              hr_treatment_dfs = 1.5,
                              hr_treatment_os = 1.3,
                              baseline_models) {
  
  if(nrow(dataset) == 0) {
    stop("OUTCOME_SIMULATION ERROR: Dataset is empty")
  }
  
  if(length(treatment) != nrow(dataset)) {
    stop("OUTCOME_SIMULATION ERROR: Treatment length does not match dataset rows")
  }
  
  if(any(is.na(treatment))) {
    stop("OUTCOME_SIMULATION ERROR: Treatment contains NA values")
  }
  
  if(!all(treatment %in% c(0, 1))) {
    stop("OUTCOME_SIMULATION ERROR: Treatment must be coded 0/1")
  }
  
  if(!is.finite(hr_treatment_dfs) || hr_treatment_dfs <= 0) {
    stop("OUTCOME_SIMULATION ERROR: Invalid hr_treatment_dfs")
  }
  
  if(!is.finite(hr_treatment_os) || hr_treatment_os <= 0) {
    stop("OUTCOME_SIMULATION ERROR: Invalid hr_treatment_os")
  }
  
  # Helper: extract coefficient, fail if invalid
  get_coef <- function(coef_vector, var_name) {
    if(is.null(coef_vector) || length(coef_vector) == 0) {
      stop(paste("OUTCOME_SIMULATION ERROR: Coefficient vector is empty for", var_name))
    }
    
    if(!(var_name %in% names(coef_vector))) {
      return(0)  # okay only if variable truly not in fitted model
    }
    
    coef_val <- coef_vector[var_name]
    
    if(is.na(coef_val) || !is.finite(coef_val)) {
      stop(paste("OUTCOME_SIMULATION ERROR: Invalid coefficient for", var_name))
    }
    
    return(as.numeric(coef_val))
  }
  
  # Initialize linear predictors with treatment effect
  linear_pred_dfs <- log(hr_treatment_dfs) * treatment
  linear_pred_os  <- log(hr_treatment_os)  * treatment
  
  # Binary covariates
  binary_vars <- intersect(
    c("Age_Dx_bin", "pT_7th_bin", "pN_7th_bin", "ENE_bin", "PORT_bin"),
    colnames(dataset)
  )
  
  for(var in binary_vars) {
    dfs_coef <- get_coef(baseline_models$dfs_coef, var)
    os_coef  <- get_coef(baseline_models$os_coef, var)
    
    linear_pred_dfs <- linear_pred_dfs + dfs_coef * dataset[[var]]
    linear_pred_os  <- linear_pred_os  + os_coef  * dataset[[var]]
  }
  
  # Continuous covariates: use RAW values to match fitted Cox model
  cont_vars <- intersect(c("Max_tumor_size", "Total_LN_positive_"), colnames(dataset))
  
  for(var in cont_vars) {
    if(any(is.na(dataset[[var]]))) {
      stop(paste("OUTCOME_SIMULATION ERROR:", var, "contains NA values"))
    }
    if(any(!is.finite(dataset[[var]]))) {
      stop(paste("OUTCOME_SIMULATION ERROR:", var, "contains non-finite values"))
    }
    
    dfs_coef <- get_coef(baseline_models$dfs_coef, var)
    os_coef  <- get_coef(baseline_models$os_coef, var)
    
    linear_pred_dfs <- linear_pred_dfs + dfs_coef * dataset[[var]]
    linear_pred_os  <- linear_pred_os  + os_coef  * dataset[[var]]
  }
  
  # Validate linear predictors
  if(any(!is.finite(linear_pred_dfs))) {
    stop("OUTCOME_SIMULATION ERROR: DFS linear predictor contains non-finite values")
  }
  
  if(any(!is.finite(linear_pred_os))) {
    stop("OUTCOME_SIMULATION ERROR: OS linear predictor contains non-finite values")
  }
  
  # Validate Weibull parameters
  required_params <- c("dfs_lambda", "dfs_gamma", "os_lambda", "os_gamma")
  missing_params <- required_params[!required_params %in% names(baseline_models)]
  if(length(missing_params) > 0) {
    stop(paste("OUTCOME_SIMULATION ERROR: Missing Weibull parameters:",
               paste(missing_params, collapse = ", ")))
  }
  
  if(!is.finite(baseline_models$dfs_lambda) || baseline_models$dfs_lambda <= 0) {
    stop("OUTCOME_SIMULATION ERROR: Invalid dfs_lambda")
  }
  if(!is.finite(baseline_models$dfs_gamma) || baseline_models$dfs_gamma <= 0) {
    stop("OUTCOME_SIMULATION ERROR: Invalid dfs_gamma")
  }
  if(!is.finite(baseline_models$os_lambda) || baseline_models$os_lambda <= 0) {
    stop("OUTCOME_SIMULATION ERROR: Invalid os_lambda")
  }
  if(!is.finite(baseline_models$os_gamma) || baseline_models$os_gamma <= 0) {
    stop("OUTCOME_SIMULATION ERROR: Invalid os_gamma")
  }
  
  # Inverse transform sampling from Weibull PH model
  u_dfs <- runif(nrow(dataset))
  u_os  <- runif(nrow(dataset))
  
  dfs_times <- (-log(u_dfs) / (baseline_models$dfs_lambda * exp(linear_pred_dfs)))^(1 / baseline_models$dfs_gamma)
  os_times  <- (-log(u_os)  / (baseline_models$os_lambda  * exp(linear_pred_os)))^(1 / baseline_models$os_gamma)
  
  if(any(!is.finite(dfs_times)) || any(dfs_times <= 0)) {
    stop("OUTCOME_SIMULATION ERROR: Invalid DFS times generated")
  }
  
  if(any(!is.finite(os_times)) || any(os_times <= 0)) {
    stop("OUTCOME_SIMULATION ERROR: Invalid OS times generated")
  }
  
  # Administrative censoring
  censor_dfs <- runif(nrow(dataset), min = 2, max = 10)
  censor_os  <- runif(nrow(dataset), min = 3, max = 15)
  
  observed_dfs_times <- pmin(dfs_times, censor_dfs)
  observed_os_times  <- pmin(os_times, censor_os)
  
  dfs_events <- as.numeric(dfs_times <= censor_dfs)
  os_events  <- as.numeric(os_times  <= censor_os)
  
  return(list(
    DFStime = observed_dfs_times,
    DFS     = dfs_events,
    OStime  = observed_os_times,
    OS      = os_events
  ))
}
```

```{r}
calculate_smd <- function(dataset, treatment_var = "Treatment", 
                         covariate_vars = NULL, 
                         weights = NULL) {
  # If no covariate variables specified, try to use available predictors
  if(is.null(covariate_vars)) {
    all_pred_vars <- c("Age_Dx_bin", "pT_7th_bin", "pN_7th_bin", "ENE_bin", "PORT_bin", "Max_tumor_size", "Total_LN_positive_")
    covariate_vars <- intersect(all_pred_vars, colnames(dataset))
  }
  
  # If no weights provided, use equal weights
  if(is.null(weights)) {
    weights <- rep(1, nrow(dataset))
  }
  # 
  if(length(weights) != nrow(dataset)) {
    stop(paste("Weight vector length mismatch:", length(weights), "vs", nrow(dataset)))
  }
  
  if(!treatment_var %in% colnames(dataset)) {
    stop(paste("Treatment variable", treatment_var, "not found in dataset"))
  }
  
  treated <- dataset[[treatment_var]] == 1
  control <- dataset[[treatment_var]] == 0
  
  if(sum(treated) < 5 || sum(control) < 5) {
    stop(paste("Insufficient sample sizes: treated =", sum(treated), ", control =", sum(control)))
  }
  smd_results <- data.frame(
    variable = covariate_vars,
    smd_unweighted = NA,
    smd_weighted = NA,
    stringsAsFactors = FALSE
  )
  # Very weird way to code this just use length(covariate_vars)
  for(i in seq_along(covariate_vars)) {
    var <- covariate_vars[i]
    
    if(!var %in% colnames(dataset)) {
      smd_results$smd_unweighted[i] <- NA
      smd_results$smd_weighted[i] <- NA
      next
    }
    
    var_data <- dataset[[var]]
    
    # Remove missing values
    complete_idx <- !is.na(var_data)
    var_complete <- var_data[complete_idx]
    treated_complete <- treated[complete_idx]
    control_complete <- control[complete_idx]
    weights_complete <- weights[complete_idx]
    
    if(sum(treated_complete) == 0 || sum(control_complete) == 0) {
      smd_results$smd_unweighted[i] <- NA
      smd_results$smd_weighted[i] <- NA
      next
    }
    
    tryCatch({
      mean_treated <- mean(var_complete[treated_complete])
      mean_control <- mean(var_complete[control_complete])
      var_treated <- var(var_complete[treated_complete])
      var_control <- var(var_complete[control_complete])
      
      # Handle zero variance case
      if(var_treated == 0 && var_control == 0) {
        smd_results$smd_unweighted[i] <- 0
      } else {
        pooled_sd <- sqrt((var_treated + var_control) / 2)
        if(pooled_sd == 0) {
          smd_results$smd_unweighted[i] <- 0
        } else {
          smd_results$smd_unweighted[i] <- (mean_treated - mean_control) / pooled_sd
        }
      }
    }, error = function(e) {
      smd_results$smd_unweighted[i] <- NA
    })
    
    tryCatch({
      weights_treated <- weights_complete[treated_complete]
      weights_control <- weights_complete[control_complete]
      
      # Weighted means
      mean_treated_wt <- sum(var_complete[treated_complete] * weights_treated) / sum(weights_treated)
      mean_control_wt <- sum(var_complete[control_complete] * weights_control) / sum(weights_control)
      
      # Weighted variances
      var_treated_wt <- sum(weights_treated * (var_complete[treated_complete] - mean_treated_wt)^2) / sum(weights_treated)
      var_control_wt <- sum(weights_control * (var_complete[control_complete] - mean_control_wt)^2) / sum(weights_control)
      
      # Handle zero variance case
      if(var_treated_wt == 0 && var_control_wt == 0) {
        smd_results$smd_weighted[i] <- 0
      } else {
        pooled_sd_wt <- sqrt((var_treated_wt + var_control_wt) / 2)
        if(pooled_sd_wt == 0) {
          smd_results$smd_weighted[i] <- 0
        } else {
          smd_results$smd_weighted[i] <- (mean_treated_wt - mean_control_wt) / pooled_sd_wt
        }
      }
    }, error = function(e) {
      smd_results$smd_weighted[i] <- NA
    })
  }
  
  return(smd_results)
}
```

```{r core_analysis_function_helpers}
# Helper functions for the three different methods
# IPTW helper
ff.att.iptw <- function(dataset, available_vars, include_diagnostics = TRUE) {
  
  balance_formula <- as.formula(paste("Treatment ~", paste(available_vars, collapse = " + ")))
  
  ws <- WeightIt::weightit(
    balance_formula,
    data = dataset,
    method = "glm",
    estimand = "ATT"
  )
  
  final_weights <- ws$weights
  
  if (include_diagnostics) {
    cat("IPTW completed\n")
    cat("Weight range:", round(min(final_weights), 4), "to", round(max(final_weights), 4), "\n")
    cat("Effective sample size:", round(sum(final_weights)), "\n")
  }
  
  return(final_weights)
}

# Entropy balancing helper
ff.att.ebal <- function(dataset, available_vars, include_diagnostics = TRUE) {
  
  # Data quality checks 
  balance_data <- dataset[complete.cases(dataset[, c("Treatment", available_vars), drop = FALSE]), ]
  
  if (nrow(balance_data) < 50)
    stop(paste("EBAL ERROR: Insufficient sample size:", nrow(balance_data), "< 50 required"))
  
  treat_dist <- table(balance_data$Treatment)
  if (min(treat_dist) < 10)
    stop(paste("EBAL ERROR: Insufficient treatment group sizes:", paste(treat_dist, collapse = ", ")))
  
  # Check for near-zero variance
  var_check <- sapply(available_vars, function(v) var(balance_data[[v]], na.rm = TRUE) > 1e-8)
  if (!all(var_check))
    stop(paste("EBAL ERROR: Near-zero variance variables:", paste(available_vars[!var_check], collapse = ", ")))
  
  # Check for rank deficiency
  X_matrix <- as.matrix(balance_data[, available_vars, drop = FALSE])
  if (qr(X_matrix)$rank < ncol(X_matrix))
    stop("EBAL ERROR: Design matrix is rank deficient")
  
  # Fit entropy balancing 
  balance_formula <- as.formula(paste("Treatment ~", paste(available_vars, collapse = " + ")))
  
  tryCatch({
    ws <- WeightIt::weightit(
      balance_formula,
      data = balance_data,
      method = "ebal",
      estimand = "ATT",
      stabilize = TRUE,
      moments = 1,
      int = FALSE,
      missing = "ind"
    )
  }, error = function(e) {
    stop(paste(
      "EBAL ERROR:", e$message,
      "\n  Sample size:", nrow(balance_data),
      "\n  Variables:", paste(available_vars, collapse = ", "),
      "\n  Treatment distribution:", paste(table(balance_data$Treatment), collapse = ", ")
    ))
  })
  
  raw_weights <- ws$weights
  # Validate
  if (any(is.na(raw_weights)))     
    stop(paste("EBAL ERROR:", sum(is.na(raw_weights)), "NA weights"))
  if (any(!is.finite(raw_weights))) 
    stop(paste("EBAL ERROR:", sum(!is.finite(raw_weights)), "non-finite weights"))
  if (any(raw_weights <= 0))        
    stop(paste("EBAL ERROR:", sum(raw_weights <= 0), "non-positive weights"))
  
  # Conservative weight trimming (only if extreme) 
  weight_ratio       <- max(raw_weights) / min(raw_weights)
  weight_range_ratio <- max(raw_weights) / mean(raw_weights)
  
  if (include_diagnostics) {
    cat("EBAL weight statistics:\n")
    cat("Range:", round(min(raw_weights), 4), "to", round(max(raw_weights), 4), "\n")
    cat("Max/Min ratio:", round(weight_ratio, 2), "\n")
  }
  
  if (weight_ratio > 100 || weight_range_ratio > 50) {
    trimmed <- pmax(quantile(raw_weights, 0.02),
                    pmin(raw_weights, quantile(raw_weights, 0.98)))
    if (any(trimmed <= 0) || any(!is.finite(trimmed)))
      stop("EBAL ERROR: Weight trimming produced invalid weights")
    cat("Trimmed", sum(raw_weights != trimmed), "extreme weights\n")
    ebal_weights <- trimmed
  } else {
    ebal_weights <- raw_weights
  }
  
  # Map weights back to full dataset (weight = 1 for incomplete cases) 
  final_weights <- rep(1, nrow(dataset))
  complete_idx  <- complete.cases(dataset[, c("Treatment", available_vars), drop = FALSE])
  final_weights[complete_idx] <- ebal_weights
  
  if (include_diagnostics) {
    cat("EBAL completed\n")
    cat("Final weight range:", round(min(final_weights), 4), "to", round(max(final_weights), 4), "\n")
    cat("Effective sample size:", round(sum(final_weights)), "\n")
  }
  
  return(final_weights)
}
# Matching helper
ff.att.match <- function(dataset, available_vars, include_diagnostics = TRUE) {
  
  # Complete cases only 
  match_vars   <- c("Treatment", available_vars)
  complete_idx <- complete.cases(dataset[, match_vars, drop = FALSE])
  match_data   <- dataset[complete_idx, ]
  
  if(nrow(match_data) < 20)
    stop(paste("MATCH ERROR: Insufficient complete cases:", nrow(match_data)))
  
  treat_dist <- table(match_data$Treatment)
  if(length(treat_dist) != 2 || min(treat_dist) < 5)
    stop(paste("MATCH ERROR: Insufficient treatment distribution:", 
               paste(treat_dist, collapse = ", ")))
  
  ps_formula <- as.formula(paste("Treatment ~", paste(available_vars, collapse = " + ")))
  
  match_out <- MatchIt::matchit(
    ps_formula,
    data = match_data,
    method = "full",  
    distance = "glm",
    estimand = "ATT"
  )
  
  # Extract weights aligned to match_data
  matched_weights <- match_out$weights
  
  # Validate — full matching should not produce zero weights
  if(any(matched_weights < 0) || any(!is.finite(matched_weights)))
    stop("MATCH ERROR: Invalid weights from full matching")
  
  # Map back to full dataset
  # Subjects excluded due to missing data get weight 0
  # coxph will subset to weight > 0 automatically
  final_weights <- rep(0, nrow(dataset))
  final_weights[complete_idx] <- matched_weights
  
  n_matched <- sum(final_weights > 0)
  if(n_matched < 10)
    stop(paste("MATCH ERROR: Too few matched subjects:", n_matched))
  
  if(include_diagnostics) {
    cat("MATCH completed\n")
    cat("Method: full matching (ATT)\n")
    cat("Matched sample size:", n_matched, "\n")
    cat("Weight range:", round(min(final_weights[final_weights > 0]), 4),
        "to", round(max(final_weights), 4), "\n")
    cat("Effective sample size:", round(sum(final_weights)), "\n")
  }
  return(final_weights)
}
```

We combine all three into the analysis function. 
```{r core_analysis_function}
ff.att.analysis <- function(dataset,
                            outcome_type = "DFS",
                            method = "iptw",
                            pred_vars = NULL,
                            include_balance = FALSE,
                            include_diagnostics = TRUE) {
  
  # Outcome variables
  if (outcome_type == "DFS") {
    time_var <- "DFStime"
    event_var <- "DFS"
  } else if (outcome_type == "OS") {
    time_var <- "OStime"
    event_var <- "OS"
  } else {
    stop("FF.ATT.ANALYSIS ERROR: outcome_type must be 'DFS' or 'OS'")
  }

  required_cols <- c("Treatment", time_var, event_var)
  missing_cols <- required_cols[!required_cols %in% colnames(dataset)]
  # Column validation 
  if (length(missing_cols) > 0) {
    print(colnames(dataset))
    stop(paste("FF.ATT.ANALYSIS ERROR: Missing columns:", paste(missing_cols, collapse = ", ")))
  }
  # Predictor selection
  all_pred_vars <- c("Age_Dx_bin", "pT_7th_bin", "pN_7th_bin", "ENE_bin",
                     "PORT_bin", "Max_tumor_size", "Total_LN_positive_")
  
  available_vars <- intersect(
    if (is.null(pred_vars)) all_pred_vars else pred_vars,
    colnames(dataset)
  )
  
  if (length(available_vars) < 2)
    stop(paste("FF.ATT.ANALYSIS ERROR: Need at least 2 predictor variables. Found:",
               paste(available_vars, collapse = ", ")))
  
  # Clean outcome rows 
  clean_idx <- !is.na(dataset$Treatment) &
               !is.na(dataset[[time_var]])  &
               !is.na(dataset[[event_var]]) &
               dataset[[time_var]] > 0      &
               dataset[[event_var]] %in% c(0, 1)
  
  if (sum(clean_idx) < 20)
    stop(paste("FF.ATT.ANALYSIS ERROR: Insufficient clean rows:", sum(clean_idx)))
  
  dataset <- dataset[clean_idx, ]
  
  treat_counts <- table(dataset$Treatment)
  if (length(treat_counts) != 2 || min(treat_counts) < 5)
    stop(paste("FF.ATT.ANALYSIS ERROR: Treatment distribution:", paste(treat_counts, collapse = ", ")))
  
  # Dispatch to weighting/matching helper 
  final_weights <- switch(method,
    iptw = ff.att.iptw( dataset, available_vars, include_diagnostics),
    ebal = ff.att.ebal( dataset, available_vars, include_diagnostics),
    match = ff.att.match(dataset, available_vars, include_diagnostics),
    stop(paste("FF.ATT.ANALYSIS ERROR: method must be 'iptw', 'ebal', or 'match'. Got:", method))
  )
  
  # Final weight validation
  if (length(final_weights) != nrow(dataset))
    stop(paste("FF.ATT.ANALYSIS ERROR: Weight length mismatch:",
               length(final_weights), "vs", nrow(dataset)))
  
  if (any(!is.finite(final_weights)))
    stop(paste("FF.ATT.ANALYSIS ERROR:", sum(!is.finite(final_weights)), "non-finite weights"))
  
  # Note: matching produces 0-weights for discarded units which is valid
  if (method != "match" && any(final_weights <= 0))
    stop(paste("FF.ATT.ANALYSIS ERROR:", sum(final_weights <= 0), "non-positive weights"))
  
  if (include_diagnostics) {
    cat("Weight validation passed |",
        "range:", round(min(final_weights), 4), "–", round(max(final_weights), 4),
        "| effective n:", round(sum(final_weights)), "\n")
  }
  analysis_data   <- dataset[final_weights > 0, ]
  analysis_weights <- final_weights[final_weights > 0]
  # Weighted Cox model 
  surv_formula <- as.formula(paste("Surv(", time_var, ",", event_var, ") ~ Treatment"))
  
  cox_model <- coxph(
    surv_formula,
    data = analysis_data,
    weights = analysis_weights,
    robust = TRUE
  )
  
  coef_val <- coef(cox_model)[1]
  var_val <- cox_model$var[1, 1]
  
  if (is.na(coef_val) || is.na(var_val) || var_val <= 0)
    stop("FF.ATT.ANALYSIS ERROR: Cox model produced invalid results")
  
  se_val <- sqrt(var_val)
  z_stat <- coef_val / se_val
  
  # Balance assessment (optional)
  balance_out <- NULL
  if (include_balance) {
    balance_data <- dataset[complete.cases(dataset[, c("Treatment", available_vars), drop = FALSE]), ]
    balance_weights <- final_weights[complete.cases(dataset[, c("Treatment", available_vars), drop = FALSE])]
    
    balance_out <- tryCatch({
      baseline_smd <- calculate_smd(balance_data, covariate_vars = available_vars)
      weighted_smd <- calculate_smd(balance_data, covariate_vars = available_vars,
                                    weights = balance_weights)
      
      smd_df <- data.frame(
        variable = baseline_smd$variable,
        smd_unweighted = baseline_smd$smd_unweighted,
        smd_weighted = weighted_smd$smd_weighted
      )
      
      list(
        good_balance_pct = mean(abs(smd_df$smd_weighted) < 0.1) * 100,
        max_abs_smd = max(abs(smd_df$smd_weighted)),
        smd_results = smd_df
      )
    }, error = function(e) NULL)
  }
  
  # Return 
  list(
    hr = exp(coef_val),
    coef = coef_val,
    se = se_val,
    pvalue = 2 * (1 - pnorm(abs(z_stat))),
    ci_lower = exp(coef_val - 1.96 * se_val),
    ci_upper = exp(coef_val + 1.96 * se_val),
    log_hr  = coef_val,
    method = method,
    weights_summary = list(
      min = min(final_weights),
      max = max(final_weights),
      mean = mean(final_weights),
      effective_n = sum(final_weights)
    ),
    balance_assessment = balance_out
  )
}
```
Let's try to simulate a dataset with what we have so far.
The workflow is:

(1) Use real OSCC data to estimate calibration parameters such as empirical distributions done using `ff.mean.sigma` and ``ff.dummy.` 

(2) Simulate covariates from the empirical distribution functions through latent multivariate Normal structures, convert binary variables. This is done through `simulate_covariates_empirical`. 

(3) Simulate treatment assignment of engraftment and PORT through a logistic regression on the covariates. This is done through `estimate_treatment_models`. 

(4) Simulate outcome (DFS/OS) and their times. First, the  survival estimates from the baseline data are estimated through `estimate_baseline_models`. Using these baseline models, we can simulate the given outcomes (events) and the survival times thourgh `simulate_outcomes`. 

(5) Assemble the full dataset

The function ``ff.simulate.empirical` will encapsulate this all. This function takes in the empirical distribution of the data we want to simulate, as well as the treatment and outcome models to simulate the data. It can be tuned to a specific HR as well. 

```{r simulate_dataset}
ff.simulate.empirical <- function(empirical_dist, n.sim, hr_treatment_dfs, hr_treatment_os, 
                                 treatment_type, variables, cat_vars, cont_vars, 
                                 baseline_models, treatment_models) {
  
  cat("Starting empirically-grounded simulation:\n")
  cat("Sample size:", n.sim, "\n")
  cat("Treatment type:", treatment_type, "\n")
  cat("DFS HR:", hr_treatment_dfs, ", OS HR:", hr_treatment_os, "\n")
  
  if(n.sim <= 0) {
    stop(paste("FF.SIMULATE.EMPIRICAL ERROR: Invalid sample size:", n.sim))
  }
  
  if(!hr_treatment_dfs > 0 || !is.finite(hr_treatment_dfs)) {
    stop(paste("FF.SIMULATE.EMPIRICAL ERROR: Invalid DFS hazard ratio:", hr_treatment_dfs))
  }
  
  if(!hr_treatment_os > 0 || !is.finite(hr_treatment_os)) {
    stop(paste("FF.SIMULATE.EMPIRICAL ERROR: Invalid OS hazard ratio:", hr_treatment_os))
  }
  
  # Simulate covariates
  sim_data <- simulate_covariates_empirical(empirical_dist, n.sim, variables, cat_vars, cont_vars)
  cat("Empirical sample created with", nrow(sim_data), "subjects\n")
  
  # Assign treatment
  sim_data$Treatment <- simulate_treatment_assignment(sim_data, treatment_type, treatment_models)
  
  # Generate outcomes
  outcomes <- simulate_outcomes(sim_data, sim_data$Treatment, 
                               hr_treatment_dfs, hr_treatment_os, baseline_models)
  
  sim_data$DFStime <- outcomes$DFStime
  sim_data$DFS <- outcomes$DFS
  sim_data$OStime <- outcomes$OStime
  sim_data$OS <- outcomes$OS
  
  # Validate final dataset
  if(any(is.na(sim_data$Treatment))) {
    stop("FF.SIMULATE.EMPIRICAL ERROR: Missing treatment assignments")
  }
  
  if(any(is.na(sim_data$DFStime)) || any(is.na(sim_data$DFS))) {
    stop("FF.SIMULATE.EMPIRICAL ERROR: Missing DFS data")
  }
  
  if(any(is.na(sim_data$OStime)) || any(is.na(sim_data$OS))) {
    stop("FF.SIMULATE.EMPIRICAL ERROR: Missing OS data")
  }
  
  # Summary statistics
  cat("Data columns:", paste(colnames(sim_data), collapse = ", "), "\n")
  cat("Treatment prevalence:", round(mean(sim_data$Treatment), 3), "\n")
  cat("DFS: median time =", round(median(sim_data$DFStime), 1), 
      ", event rate =", round(mean(sim_data$DFS), 3), "\n")
  cat("OS: median time =", round(median(sim_data$OStime), 1), 
      ", event rate =", round(mean(sim_data$OS), 3), "\n")
  cat("Dataset generated with", nrow(sim_data), "observations\n\n")
  
  return(sim_data)
}
```

# Missing Data Simulation

There are three standard missing data mechanisms, defined by what the probability of missingness depends on:

(1) MCAR — Missing Completely At Random
Missingness is purely random, unrelated to anything in the data. The probability of being missing is the same for everyone regardless of their observed or unobserved values. This is the strongest and most unrealistic assumption — it essentially means the complete cases are a random subsample of the full data, so naive complete case analysis is still unbiased, just less efficient. The original simulation was only on MCAR. 

(2) MAR — Missing At Random
Missingness depends on observed data but not on the missing values themselves. Example: older patients are less likely to have LN positivity recorded, but conditional on age, missingness doesn't depend on the actual LN value. This is the most commonly assumed mechanism in practice and is what most modern methods (multiple imputation, IPW for missing data) target. "At random" is a misnomer as it just means the missingness is explainable by what you've observed.

Mathematically this is seen as 
$$P (M = 1|(y_{mis}, y_{obs}) ) = P (M = 1|y_{obs}) $$

(3) MNAR — Missing Not At Random
Missingness depends on the unobserved value itself. Example: patients with very high LN positivity are less likely to have it recorded precisely because it's so high. This is the most dangerous and hardest to handle — no standard method can correct for it without additional assumptions, because the information about why data is missing is contained in the missing data itself.

The number matters as well - how many variables do we want missing? To begin let's just focus on the covariates first before moving on to treatment/outcome being missing. We can look at three options: single variable missing, two missing variables and all missing.

Next we can look at how many observations will be missing in each variable. April's paper looked at 0%, 25% and 50%. I think it's clear that if we have half our data missing this is a major problem to begin with. Instead it might be easier to look at smaller numbers first - 0%, 10% and 25%. 

## How to Simulate Missingness

(1) MCAR is the easiest - it's purely random. In a given dataset we can just remove any number of observations purely at random. This is done with the `sample` function with replacement off. This randomly samples from the data without repeats - which will be the ones removed/replace with NA. 

The following is the function to do this and an example of it.

```{r mcar_simulate}
ff.miss.sub.mcar <- function(dataset, in.subj, in.pred) {
  if(in.subj == 0) return(dataset)
  
  if(in.subj < 0 || in.subj > 1) {
    stop(paste("FF.MISS.SUB.MCAR ERROR: Invalid missingness proportion:", in.subj))
  }
  
  if(length(in.pred) == 0) {
    stop("FF.MISS.SUB.MCAR ERROR: No predictor variables specified for missingness")
  }
  
  missing_vars <- in.pred[!in.pred %in% colnames(dataset)]
  if(length(missing_vars) > 0) {
    stop(paste("FF.MISS.SUB.VAR ERROR: Variables not found in dataset:", paste(missing_vars, collapse = ", ")))
  }
  
  n_total <- nrow(dataset)
  n_missing <- round(n_total * in.subj)
  
  if(n_missing >= n_total) {
    stop(paste("FF.MISS.SUB.MCAR ERROR: Would make all subjects missing:", n_missing, ">=", n_total))
  }
  
  # Randomly select subjects for missing data
  missing_subjects <- sample(1:n_total, n_missing, replace = FALSE)
  
  # Apply missingness to specified predictors
  for(var in in.pred) {
    dataset[missing_subjects, var] <- NA
  }
  
  cat("Applied", round(in.subj*100, 1), "% missingness to", length(in.pred), "variables\n")
  return(dataset)
}
```

(2) MAR - the challenge is determining what clinically sensible choices for missing data would look like. There are a few steps we need to take.
First - determine what makes sense for MAR data to be. 
Second - make the functions that actually simulate MAR.
Third - develop tools to assess whether MAR was actually achieved.

MAR is merely the probability of a value being missing depending on observed data. For example, older patients are less likely to have lab values recorded. The probability of missingness in a variable $Y$ depends on an observed auxiliary variable $X$ but not on the values of $Y$ itself. 

Some examples of MAR scenarios we might be interested in. The most common is when missingness is dependent on baseline covariates:
(1) Age at diagnosis (Age_Dx)
(2) Tumor stage (pT_7th, pN_7th)
(3) ENE status
(4) Treatment (engraftment / PORT)

Next we might have missingness on treatment assignment. Those are being treated likely will experience less drop-out and have more follow up. Vice-versa untreated individuals might have more reasons to drop out of a study and hence have missing value.

We can also have outcome-dependent MAR - missingness may be due to outcomes. 



```{r mar_mnar_simulate}
ff.miss.mar <- function(dataset, in.subj, in.pred, drivers) {
  # drivers: named list of which observed variables drive missingness
  # examples: drivers = list("ENE_bin" = c("Total_LN_positive_", "pN_7th_bin"))
  
  if(in.subj == 0) return(dataset)
  
  n_total <- nrow(dataset)
  
  for(var in in.pred) {
    
    # Get the variables that drive missingness for this variable
    driver_vars <- drivers[[var]]
    
    if(is.null(driver_vars)) {
      stop(paste("FF.MISS.MAR ERROR: No drivers specified for", var))
    }

    if(var %in% driver_vars) {
      stop(paste("FF.MISS.MAR ERROR: Target variable cannot be its own driver:", var))
    }
    
    missing_vars <- driver_vars[!driver_vars %in% colnames(dataset)]
    if(length(missing_vars) > 0) {
      stop(paste("FF.MISS.MAR ERROR: Driver variables not found:", 
                 paste(missing_vars, collapse = ", ")))
    }
    
    # Build linear combination of driver variables
    lp <- rowSums(scale(dataset[, driver_vars, drop = FALSE]))
    
    # Calibrate intercept to achieve desired missingness proportion
    # This make sure we get the exact proportion
    # Use uniroot to find intercept that gives mean(prob) = in.subj
    intercept <- uniroot(function(a) {
      mean(plogis(a + lp)) - in.subj
    }, interval = c(-20, 20))$root
    
    # Compute missingness probability for each subject
    miss_prob <- plogis(intercept + lp)
    
    # Draw missingness indicator
    missing_subjects <- rbinom(n_total, 1, miss_prob) == 1
    
    dataset[missing_subjects, var] <- NA
    
    cat("Applied MAR missingness to", var, 
        "- actual proportion missing:", round(mean(missing_subjects), 3), "\n")
  }
  
  return(dataset)
}

# MNAR: missingness probability depends on the VARIABLE ITSELF

ff.miss.mnar <- function(dataset, in.subj, in.pred) {
  if (in.subj == 0) return(dataset)
  
  if (in.subj < 0 || in.subj > 1)
    stop("FF.MISS.MNAR ERROR: Invalid missingness proportion")
  
  n_total <- nrow(dataset)
  
  for (var in in.pred) {
    
    if (!var %in% colnames(dataset))
      stop(paste("FF.MISS.MNAR ERROR: Variable not found:", var))
    if (!is.numeric(dataset[[var]]))
      stop(paste("FF.MISS.MNAR ERROR: Variable must be numeric:", var))
    
    var_vals <- dataset[[var]]
    
    if (all(is.na(var_vals)))
      stop(paste("FF.MISS.MNAR ERROR: Variable is entirely NA:", var))
    
    # Standardize variable using scale() — consistent with MAR approach
    var_scaled <- scale(var_vals)
    
    # Calibrate intercept to achieve exact target proportion
    intercept <- uniroot(function(a) {
      mean(plogis(a + var_scaled)) - in.subj
    }, interval = c(-20, 20))$root
    
    # Compute missingness probability for each subject
    miss_prob <- plogis(intercept + var_scaled)
    
    # Draw missingness indicator
    missing_subjects <- rbinom(n_total, 1, miss_prob) == 1
    
    dataset[missing_subjects, var] <- NA
    
    cat("MNAR: Applied", round(mean(missing_subjects) * 100, 1),
        "% missingness to", var,
        "- actual proportion missing:", round(mean(missing_subjects), 3), "\n")
  }
  return(dataset)
}
```

Using these three functions we now in our simulation framework we can generate datasets based on different forms of missingness. The following function will do just that. The `Generate` function has two versions that both work. The first generates/simulate a full dataset and then simulates three additional on top that implement the three missingness types - thus simualting a total of four datasets. The second version later on is an alternative version as a back-up which only does the initial complete dataset (and missingness is simulated later on top of this). 

```{r generate}
Generate <- function(condition, fixed_objects = NULL) {
  library(survival)
  library(MASS)
  library(WeightIt)
  library(ebal)
  
  if(is.null(fixed_objects)) {
    stop("GENERATE ERROR: fixed_objects is NULL")
  }
  
  required_objects <- c("pred.vars.processed", "cat.vars.processed", "cont.vars.processed",
                         "empirical_dist", "baseline_models", "treatment_models", "mar_drivers")
  missing_objects <- required_objects[!required_objects %in% names(fixed_objects)]
  if(length(missing_objects) > 0) {
    stop(paste("GENERATE ERROR: Missing fixed objects:", paste(missing_objects, collapse = ", ")))
  }
  
  pred_vars <- fixed_objects$pred.vars.processed
  cat_vars <- fixed_objects$cat.vars.processed
  cont_vars <- fixed_objects$cont.vars.processed
  
  required_conditions <- c("input_ss", "input_hr_dfs", "input_hr_os", 
                            "input_mis_subj", "treatment_type", "miss_vars")
  missing_conditions <- required_conditions[!required_conditions %in% names(condition)]
  if(length(missing_conditions) > 0) {
    stop(paste("GENERATE ERROR: Missing condition parameters:", 
               paste(missing_conditions, collapse = ", ")))
  }
  
  n.sim <- condition$input_ss
  hr.dfs <- condition$input_hr_dfs
  hr.os <- condition$input_hr_os
  in.mis.subj <- condition$input_mis_subj
  treatment_type <- condition$treatment_type
  miss_vars <- condition$miss_vars
  
  # Fixes miss_vars
  if(is.list(miss_vars)) miss_vars <- unlist(miss_vars)
  if(is.character(miss_vars) && length(miss_vars) == 1) {
    miss_vars <- strsplit(miss_vars, ",")[[1]]
    miss_vars <- trimws(miss_vars) 
}
  # Simulate base dataset
  sim.dataset <- ff.simulate.empirical(
    empirical_dist = fixed_objects$empirical_dist,
    n.sim = n.sim,
    hr_treatment_dfs = hr.dfs,
    hr_treatment_os = hr.os,
    treatment_type = treatment_type,
    variables = pred_vars,
    cat_vars = cat_vars,
    cont_vars = cont_vars,
    baseline_models = fixed_objects$baseline_models,
    treatment_models = fixed_objects$treatment_models
  )
  
  # Dataset 1: No missingness 
  dataset.complete <- sim.dataset
  
  if(in.mis.subj > 0) {
    
    # Validate miss_vars exist in dataset
    missing_vars <- miss_vars[!miss_vars %in% colnames(sim.dataset)]
    if(length(missing_vars) > 0) {
      stop(paste("GENERATE ERROR: miss_vars not found in dataset:", 
                 paste(missing_vars, collapse = ", ")))
    }
    
    # Dataset 2: MCAR 
    dataset.mcar <- ff.miss.sub.mcar(
      dataset = sim.dataset,
      in.subj = in.mis.subj,
      in.pred = miss_vars
    )
    
    # Dataset 3: MAR 
    # Validate that all miss_vars have drivers defined
    missing_drivers <- miss_vars[!miss_vars %in% names(fixed_objects$mar_drivers)]
    if(length(missing_drivers) > 0) {
      stop(paste("GENERATE ERROR: No MAR drivers defined for:", 
                 paste(missing_drivers, collapse = ", ")))
    }
    
    dataset.mar <- ff.miss.mar(
      dataset = sim.dataset,
      in.subj = in.mis.subj,
      in.pred = miss_vars,
      drivers = fixed_objects$mar_drivers[miss_vars]
    )
    
    # Dataset 4: MNAR
    dataset.mnar <- ff.miss.mnar(
      dataset = sim.dataset,
      in.subj = in.mis.subj,
      in.pred = miss_vars
    )
    
  } else {
    # No missingness — all datasets identical to complete data
    dataset.mcar <- sim.dataset
    dataset.mar <- sim.dataset
    dataset.mnar <- sim.dataset
  }
  return(list(
    dataset.complete = dataset.complete,
    dataset.mcar = dataset.mcar,
    dataset.mar = dataset.mar,
    dataset.mnar = dataset.mnar
  ))
}
```

```{r generate_2}
Generate2 <- function(condition, fixed_objects = NULL) {
  library(survival)
  library(MASS)
  library(WeightIt)
  library(ebal)
  
  if(is.null(fixed_objects)) {
    stop("GENERATE ERROR: fixed_objects is NULL")
  }
  
  required_objects <- c("pred.vars.processed", "cat.vars.processed", "cont.vars.processed",
                         "empirical_dist", "baseline_models", "treatment_models", "mar_drivers")
  missing_objects <- required_objects[!required_objects %in% names(fixed_objects)]
  if(length(missing_objects) > 0) {
    stop(paste("GENERATE ERROR: Missing fixed objects:", paste(missing_objects, collapse = ", ")))
  }
  
  pred_vars <- fixed_objects$pred.vars.processed
  cat_vars <- fixed_objects$cat.vars.processed
  cont_vars <- fixed_objects$cont.vars.processed
  
  required_conditions <- c("input_ss", "input_hr_dfs", "input_hr_os", 
                            "input_mis_subj", "treatment_type", "miss_vars")
  missing_conditions <- required_conditions[!required_conditions %in% names(condition)]
  if(length(missing_conditions) > 0) {
    stop(paste("GENERATE ERROR: Missing condition parameters:", 
               paste(missing_conditions, collapse = ", ")))
  }
  
  n.sim <- condition$input_ss
  hr.dfs <- condition$input_hr_dfs
  hr.os  <- condition$input_hr_os
  in.mis.subj <- condition$input_mis_subj
  treatment_type <- condition$treatment_type
  miss_vars <- condition$miss_vars
  
  if(is.list(miss_vars)) miss_vars <- unlist(miss_vars)
  # --- Simulate base dataset ---
  sim.dataset <- ff.simulate.empirical(
    empirical_dist = fixed_objects$empirical_dist,
    n.sim = n.sim,
    hr_treatment_dfs = hr.dfs,
    hr_treatment_os = hr.os,
    treatment_type = treatment_type,
    variables = pred_vars,
    cat_vars = cat_vars,
    cont_vars = cont_vars,
    baseline_models = fixed_objects$baseline_models,
    treatment_models = fixed_objects$treatment_models
  )
  
  # Dataset 1: No missingness 
  dataset.complete <- sim.dataset
  
  return(dataset.complete)
}
```

# Running the overall simulation

The library SimDesign is used for the overall simulation. It provides a framework for implementing simulation studies. It comprises of three parts: generate, analysis and summarise. Generate is given above - it generates datasets. Analysis is where we use our methods on the simulations. In this case it will be running the three causal methods on each study (as well as deciding whether we run multiple imputation or not). Summarise just summarise the results of each simulation condition over all replications. 

The following is a helper function for the analysis function which implements the complete case vs. multiple imputation option and performs MI. As mentioned the analysis function will when given a dataset with or without missingness, check whether we want complete case or multiple imputation before running the specified causal method. 
```{r mi_analysis}
enhanced_parallel_mi_analysis <- function(dataset, pred_vars, cat_vars, cont_vars,
                                          causal_method, in.mis.subj, nSim,
                                          sim_seed = NULL) {
  suppressPackageStartupMessages({
    library(survival)
    library(mice)
    library(purrr)
    library(WeightIt)
    library(ebal)
    library(dplyr)
  })
  
  cat("=== ENHANCED PARALLEL MI ANALYSIS ===\n")
  cat("Sample size:", nrow(dataset), "| Missing level:", in.mis.subj * 100, "%\n")
  cat("Imputations:", nSim, "| Method:", causal_method, "\n")
  
  # Step 1: Prepare data for MI 
  mi_data <- dataset
  original_numeric_values <- list()
  
  # Convert categorical variables to factors for mice
  for (var in cat_vars) {
    if (var %in% colnames(mi_data)) {
      original_numeric_values[[var]] <- mi_data[, var]
      if (is.numeric(mi_data[, var])) {
        unique_vals <- sort(unique(mi_data[, var][!is.na(mi_data[, var])]))
        mi_data[, var] <- factor(mi_data[, var], levels = unique_vals)
      } else {
        mi_data[, var] <- as.factor(mi_data[, var])
      }
    }
  }
  
  # Handle Treatment variable
  original_numeric_values[["Treatment"]] <- mi_data$Treatment
  if (is.numeric(mi_data$Treatment)) {
    unique_treatment_vals <- sort(unique(mi_data$Treatment[!is.na(mi_data$Treatment)]))
    mi_data$Treatment <- factor(mi_data$Treatment, levels = unique_treatment_vals)
  } else {
    mi_data$Treatment <- as.factor(mi_data$Treatment)
  }
  
  # Get variables that actually need imputation
  vars_to_impute <- pred_vars[pred_vars %in% colnames(mi_data)]
  vars_with_missing <- sapply(vars_to_impute, function(x) any(is.na(mi_data[, x])))
  actual_vars_to_impute <- vars_to_impute[vars_with_missing]
  
  if (length(actual_vars_to_impute) == 0) {
    cat("No variables need imputation - returning NULL for complete case fallback\n")
    return(NULL)
  }
  
  cat("Variables to impute:", paste(actual_vars_to_impute, collapse = ", "), "\n")
  
  # Step 2: Build imputation model 
  # Excluding outcomes causes attenuation bias toward null
  all_imp_vars <- unique(c(
    actual_vars_to_impute,
    pred_vars,
    "Treatment",
    "OStime", "OS", "DFStime", "DFS"
  ))
  all_imp_vars <- intersect(all_imp_vars, colnames(mi_data))
  
  start_imputation_time <- Sys.time()
  
  # CRITICAL FIX: use maxit=0 dry run to build predictor matrix
  # avoids running mice twice unnecessarily
  init <- mice::mice(
    mi_data[, all_imp_vars, drop = FALSE],
    maxit = 0,
    printFlag = FALSE
  )
  
  pred_matrix <- init$predictorMatrix
  
  # FE-MI: set treatment as fixed effect in imputation model for all missing vars
  pred_matrix[actual_vars_to_impute, "Treatment"] <- 2
  
  # Prevent outcomes from being imputed (they are complete)
  # but keep them as predictors in the imputation model
  outcome_vars <- intersect(c("OStime", "OS", "DFStime", "DFS"), colnames(mi_data))
  for (ov in outcome_vars) {
    if (ov %in% rownames(pred_matrix)) {
      pred_matrix[ov, ] <- 0  # do not impute outcomes
    }
  }
  
  # CRITICAL FIX: dynamic seed derived from simulation replicate
  # fixed seed = identical imputations across all replicates
  imp_seed <- if (!is.null(sim_seed)) sim_seed else NA
  
  tempData <- mice::mice(
    mi_data[, all_imp_vars, drop = FALSE],
    m = nSim,
    method = "pmm",
    predictorMatrix = pred_matrix,
    maxit = 10,
    printFlag = FALSE,
    seed = imp_seed
  )
  
  imputation_time <- as.numeric(Sys.time() - start_imputation_time)
  cat("FE-MI imputation completed in", round(imputation_time, 2), "seconds\n")
  
  # --- Step 3: Analyze each imputed dataset ---
  cat("Starting analysis of", nSim, "imputed datasets...\n")
  analysis_start_time <- Sys.time()
  
  analyze_single_imputation <- function(i, tempData, mi_data, actual_vars_to_impute,
                                        original_numeric_values, causal_method, pred_vars) {
    library(survival)
    library(WeightIt)
    library(ebal)
    
    completed_data <- mice::complete(tempData, i)
    
    # Reconstruct full dataset with imputed values
    full_data <- mi_data
    full_data[, actual_vars_to_impute] <- completed_data[, actual_vars_to_impute]
    
    # Convert all variables back to numeric
    for (var in names(original_numeric_values)) {
      if (var %in% colnames(full_data)) {
        full_data[, var] <- as.numeric(as.character(full_data[, var]))
      }
    }
    print("Multiple imputation ATT")
    tryCatch({
      att.dfs <- ff.att.analysis(full_data, outcome_type = "DFS",
                                 method = causal_method, pred_vars = pred_vars,
                                 include_balance = FALSE, include_diagnostics = FALSE)
      att.os  <- ff.att.analysis(full_data, outcome_type = "OS",
                                 method = causal_method, pred_vars = pred_vars,
                                 include_balance = FALSE, include_diagnostics = FALSE)
      list(
        att.log.hr.dfs = att.dfs$log_hr,
        att.se.dfs = att.dfs$se,
        att.log.hr.os = att.os$log_hr,
        att.se.os = att.os$se,
        success = TRUE,
        imputation_id = i
      )
    }, error = function(e) {
      list(
        att.log.hr.dfs = NA, att.se.dfs = NA,
        att.log.hr.os = NA, att.se.os  = NA,
        success = FALSE,
        error = e$message,
        imputation_id = i
      )
    })
  }
  
  MI.results <- lapply(
    X = 1:nSim,
    FUN = analyze_single_imputation,
    tempData = tempData,
    mi_data = mi_data,
    actual_vars_to_impute = actual_vars_to_impute,
    original_numeric_values = original_numeric_values,
    causal_method = causal_method,
    pred_vars = pred_vars
  )
  
  analysis_time <- as.numeric(Sys.time() - analysis_start_time)
  cat("Parallel analysis completed in", round(analysis_time, 2), "seconds\n")
  
  # Step 4: Validate results 
  successful_results <- MI.results[sapply(MI.results, function(x) x$success)]
  
  if (length(successful_results) < nSim * 0.7) {
    stop(paste("MI ERROR: Too many failed imputations:",
               length(successful_results), "/", nSim))
  }
  cat("Successful imputations:", length(successful_results), "/", nSim, "\n")
  
  # Step 5: Effective sample size
  effective_n <- round(mean(sapply(1:nSim, function(i) {
    completed_data <- mice::complete(tempData, i)
    sum(complete.cases(completed_data))
  })))
  cat("Effective sample size:", effective_n, "\n")
  
  # Step 6: Extract estimates
  dfs_loghr_list <- lapply(successful_results, function(x) x$att.log.hr.dfs)
  dfs_se_list    <- lapply(successful_results, function(x) x$att.se.dfs)
  os_loghr_list  <- lapply(successful_results, function(x) x$att.log.hr.os)
  os_se_list     <- lapply(successful_results, function(x) x$att.se.os)
  
  # Step 7: Pool using Rubin's rules
  pool_results <- tryCatch({
    
    dfs_fits <- purrr::map2(dfs_loghr_list, dfs_se_list, function(est, se) {
      if (is.na(est) || is.na(se) || se <= 0) return(NULL)
      list(
        coefficients = est,
        vcov = se^2,
        df.residual = max(5, effective_n - length(pred_vars))
      )
    })
    
    os_fits <- purrr::map2(os_loghr_list, os_se_list, function(est, se) {
      if (is.na(est) || is.na(se) || se <= 0) return(NULL)
      list(
        coefficients = est,
        vcov = se^2,
        df.residual = max(5, effective_n - length(pred_vars))
      )
    })
    
    dfs_fits <- dfs_fits[!sapply(dfs_fits, is.null)]
    os_fits <- os_fits[!sapply(os_fits, is.null)]
    
    if (length(dfs_fits) < 2 || length(os_fits) < 2) {
      stop("Too few successful imputations for pooling")
    }
    
    pooled_dfs <- mice::pool(dfs_fits)
    pooled_os <- mice::pool(os_fits)
    
    dfs_summary <- summary(pooled_dfs, conf.int = TRUE)
    os_summary <- summary(pooled_os,  conf.int = TRUE)
    
    list(
      success = TRUE,
      aux.dfs = list(
        qbar = dfs_summary$estimate[1],
        se = dfs_summary$std.error[1],
        pval = dfs_summary$p.value[1],
        ci_lower = dfs_summary$`2.5 %`[1],
        ci_upper = dfs_summary$`97.5 %`[1],
        df = dfs_summary$df[1],
        lambda = pooled_dfs$pooled$lambda[1],
        m = length(dfs_fits)
      ),
      aux.os = list(
        qbar = os_summary$estimate[1],
        se = os_summary$std.error[1],
        pval = os_summary$p.value[1],
        ci_lower = os_summary$`2.5 %`[1],
        ci_upper = os_summary$`97.5 %`[1],
        df = os_summary$df[1],
        lambda = pooled_os$pooled$lambda[1],
        m = length(os_fits)
      )
    )
    
  }, error = function(e) {
    cat("mice::pool() failed:", e$message, "- falling back to manual pooling\n")
    list(success = FALSE, error = e$message)
  })
  
  # Fallback to manual pooling if mice::pool() failed
  if (!pool_results$success) {
    cat("Using manual pooling fallback...\n")
    aux.dfs <- manual_pool_mi(dfs_loghr_list, dfs_se_list, n_obs = effective_n)
    aux.os <- manual_pool_mi(os_loghr_list,  os_se_list,  n_obs = effective_n)
  } else {
    aux.dfs <- pool_results$aux.dfs
    aux.os <- pool_results$aux.os
  }
  
  if (is.null(aux.dfs) || is.null(aux.os)) {
    stop("Both mice::pool() and manual pooling failed")
  }
  
  total_time <- imputation_time + analysis_time
  cat("MI completed | Total time:", round(total_time, 2), "seconds\n")
  cat("DFS pooled log HR:", round(aux.dfs$qbar, 3), "SE:", round(aux.dfs$se, 3),
      "lambda:", round(aux.dfs$lambda, 3), "\n")
  cat("OS  pooled log HR:", round(aux.os$qbar,  3), "SE:", round(aux.os$se,  3),
      "lambda:", round(aux.os$lambda,  3), "\n")
  
  # --- Step 8: Return pooled results ---
  return(list(
    att.pooled.dfs = exp(aux.dfs$qbar),
    att.pooled.os = exp(aux.os$qbar),
    att.se.dfs = aux.dfs$se,
    att.se.os = aux.os$se,
    att.pvalue.dfs = aux.dfs$pval,
    att.pvalue.os = aux.os$pval,
    ci_lower_dfs = aux.dfs$ci_lower,
    ci_upper_dfs = aux.dfs$ci_upper,
    ci_lower_os = aux.os$ci_lower,
    ci_upper_os = aux.os$ci_upper,
    lambda_dfs = aux.dfs$lambda,
    lambda_os  = aux.os$lambda,
    successful_imputations = length(successful_results),
    total_time_seconds = total_time
  ))
}

# The first version is broken and I don't know how to fix so we use this instead for now
enhanced_mi_analysis <- function(dataset, pred_vars, cat_vars, cont_vars,
                                 causal_method, in.mis.subj, nSim,
                                 sim_seed = NULL) {
  suppressPackageStartupMessages({
    library(survival)
    library(mice)
    library(WeightIt)
    library(ebal)
    library(dplyr)
  })
  
  # Step 1: Prepare data 
  mi_data <- dataset
  original_numeric_values <- list()
  
  for(var in cat_vars) {
    if(var %in% colnames(mi_data)) {
      original_numeric_values[[var]] <- mi_data[, var]
      mi_data[, var] <- factor(mi_data[, var])
    }
  }
  
  original_numeric_values[["Treatment"]] <- mi_data$Treatment
  mi_data$Treatment <- factor(mi_data$Treatment)
  
  # Variables that need imputation
  vars_to_impute <- pred_vars[pred_vars %in% colnames(mi_data)]
  vars_with_missing <- sapply(vars_to_impute, function(x) any(is.na(mi_data[, x])))
  actual_vars_to_impute <- vars_to_impute[vars_with_missing]
  
  if(length(actual_vars_to_impute) == 0) {
    cat("No variables need imputation\n")
    return(NULL)
  }
  
  # Step 2: Build imputation model
  # Include outcomes as predictors but not as targets
  all_imp_vars <- unique(c(
    actual_vars_to_impute, pred_vars,
    "Treatment", "OStime", "OS", "DFStime", "DFS"
  ))
  all_imp_vars <- intersect(all_imp_vars, colnames(mi_data))
  
  start_time <- Sys.time()
  
  # Dry run to get predictor matrix
  init <- mice::mice(mi_data[, all_imp_vars, drop = FALSE],
                            maxit = 0, printFlag = FALSE)
  pred_matrix <- init$predictorMatrix
  
  # FE-MI: treatment as fixed effect
  pred_matrix[actual_vars_to_impute, "Treatment"] <- 2
  
  # Block outcomes from being imputed
  outcome_vars <- intersect(c("OStime", "OS", "DFStime", "DFS"), rownames(pred_matrix))
  pred_matrix[outcome_vars, ] <- 0
  
  imp_seed <- if(!is.null(sim_seed)) sim_seed else NA
  
  tempData <- mice::mice(
    mi_data[, all_imp_vars, drop = FALSE],
    m = nSim,
    method = "pmm",
    predictorMatrix = pred_matrix,
    maxit = 10,
    printFlag = FALSE,
    seed = imp_seed
  )
  
  imputation_time <- as.numeric(Sys.time() - start_time)
  
  # Step 3: Analyse each imputed dataset sequentially
  # Sequential — avoids nested parallelism and overheating
  analysis_start <- Sys.time()
  
  MI.results <- lapply(1:nSim, function(i) {
    
    completed_data <- mice::complete(tempData, i)
    full_data <- mi_data
    full_data[, actual_vars_to_impute] <- completed_data[, actual_vars_to_impute]
    
    # Convert back to numeric
    for(var in names(original_numeric_values)) {
      if(var %in% colnames(full_data)) {
        full_data[, var] <- as.numeric(as.character(full_data[, var]))
      }
    }
    
    tryCatch({
      att.dfs <- ff.att.analysis(full_data, 
                                 outcome_type = "DFS",
                                 method = causal_method,
                                 pred_vars = pred_vars,
                                 include_balance = FALSE,
                                 include_diagnostics = FALSE)
      att.os <- ff.att.analysis(full_data, 
                                outcome_type = "OS",
                                method = causal_method,
                                pred_vars = pred_vars,
                                include_balance = FALSE,
                                include_diagnostics = FALSE)
      list(
        log_hr_dfs = att.dfs$log_hr, se_dfs = att.dfs$se,
        log_hr_os = att.os$log_hr,  se_os  = att.os$se,
        success = TRUE
      )
    }, error = function(e) {
      list(log_hr_dfs = NA, se_dfs = NA,
           log_hr_os = NA, se_os  = NA,
           success = FALSE, error = e$message)
    })
  })
  
  analysis_time <- as.numeric(Sys.time() - analysis_start)
  
  # Step 4: Validate 
  successful <- MI.results[sapply(MI.results, function(x) x$success)]
  
  if(length(successful) < nSim * 0.7)
    stop(paste("MI ERROR: Too many failures:", length(successful), "/", nSim))
  
  # Step 5: Pool using Rubin's rules 
  pool_rubin <- function(estimates, ses, n_obs) {
    
    estimates <- unlist(estimates)
    ses <- unlist(ses)
    valid <- is.finite(estimates) & is.finite(ses) & ses > 0
    estimates <- estimates[valid]
    ses <- ses[valid]
    m <- length(estimates)
    
    if(m < 2) stop("Need at least 2 valid imputations")
    
    qbar <- mean(estimates)
    Ubar <- mean(ses^2)
    B <- var(estimates)
    T_var <- Ubar + (1 + 1/m) * B
    se <- sqrt(T_var)
    lambda <- (1 + 1/m) * B / T_var
    
    # Degrees of freedom
    df_old <- (m - 1) / lambda^2
    n_parms <- length(pred_vars)
    df_obs <- ((n_obs - n_parms + 1) / (n_obs - n_parms + 3)) *
               (n_obs - n_parms) * (1 - lambda)
    df <- df_old * df_obs / (df_old + df_obs)
    df <- max(df, 3)   # floor to avoid degenerate t
    
    t_crit <- qt(0.975, df = df)
    list(
      qbar = qbar,
      se = se,
      ci_lower = qbar - t_crit * se,
      ci_upper = qbar + t_crit * se,
      pval = 2 * pt(abs(qbar / se), df = df, lower.tail = FALSE),
      df = df,
      lambda = lambda,
      m = m
    )
  }
  
  n_obs <- nrow(dataset)
  
  dfs <- pool_rubin(
    lapply(successful, function(x) x$log_hr_dfs),
    lapply(successful, function(x) x$se_dfs),
    n_obs
  )
  os <- pool_rubin(
    lapply(successful, function(x) x$log_hr_os),
    lapply(successful, function(x) x$se_os),
    n_obs
  )
  
  total_time <- imputation_time + analysis_time
  cat("MI done | Time:", round(total_time, 2), "s |",
      "DFS HR:", round(exp(dfs$qbar), 3),
      "OS HR:",  round(exp(os$qbar),  3),
      "lambda DFS:", round(dfs$lambda, 3),
      "lambda OS:",  round(os$lambda,  3), "\n")
  
  # Step 6: Return
  list(
    att.pooled.dfs = exp(dfs$qbar),
    att.pooled.os = exp(os$qbar),
    att.se.dfs = dfs$se,
    att.se.os = os$se,
    att.pvalue.dfs = dfs$pval,
    att.pvalue.os = os$pval,
    ci_lower_dfs = dfs$ci_lower,
    ci_upper_dfs = dfs$ci_upper,
    ci_lower_os = os$ci_lower,
    ci_upper_os = os$ci_upper,
    lambda_dfs = dfs$lambda,
    lambda_os = os$lambda,
    successful_imputations = length(successful),
    total_time_seconds = total_time
  )
}

# Manual implementation of Rubin's rules for MI pooling
# Manual Rubin's rules pooling — replace the entire pool_results tryCatch block
manual_pool_mi <- function(estimates, ses, n_obs) {
  
  estimates <- unlist(estimates)
  ses <- unlist(ses)
  
  # Remove invalid entries
  valid <- is.finite(estimates) & is.finite(ses) & ses > 0
  estimates <- estimates[valid]
  ses <- ses[valid]
  m <- length(estimates)
  
  if(m < 2) stop("Need at least 2 valid imputations to pool")
  
  # Rubin's rules
  qbar <- mean(estimates)                          # pooled estimate
  Ubar <- mean(ses^2)                              # within-imputation variance
  B <- var(estimates)                           # between-imputation variance
  T_var <- Ubar + (1 + 1/m) * B                    # total variance
  se    <- sqrt(T_var)                              # pooled SE
  
  # Degrees of freedom (Barnard & Rubin 1999)
  lambda <- (1 + 1/m) * B / T_var                  # fraction of missing info
  df_old <- (m - 1) / lambda^2
  df_obs <- (n_obs - length(pred_vars) + 1) / 
            (n_obs - length(pred_vars) + 3) * 
            (n_obs - length(pred_vars)) * (1 - lambda)
  df <- df_old * df_obs / (df_old + df_obs)
  
  # CI and p-value
  t_crit <- qt(0.975, df = df)
  ci_lower <- qbar - t_crit * se
  ci_upper <- qbar + t_crit * se
  pval <- 2 * pt(abs(qbar / se), df = df, lower.tail = FALSE)
  
  list(
    qbar = qbar,
    sev= se,
    ci_lower = ci_lower,
    ci_upper = ci_upper,
    pval = pval,
    df = df,
    lambda = lambda,
    m = m
  )
}
```

Here we need to look at the overall simulation framework. This simulation has three layers to consider:

Layer 1 - Data Generating Conditions

For our dataset we need to consider what conditions are we going to be simulating under. This means sample sizes and hazard ratios (for both overall survival and disease free survival).

Layer 2 - Missingness Conditions (what kind of missing data are we looking at)
This includes:
 - Which variables have missingness
 - Proportion missing (0%, 25%, 50%)
 - Mechanism (MCAR, MAR, MNAR)
 - Mechanism-specific parameters (drivers for MAR, direction for MNAR)
 
 Layer 3 — Analysis methods (how we handle the missing data and estimate the treatment effect)

 - Complete case analysis
 - Multiple imputation
 - IPTW, matching, ebal — each applied to each dataset
 
 Due to the high number of variables to consider missginess it might be best pick 2-3 meaningful pairs rather than exhaust all possible combinations. 
 When it comes to MCAR data there are no extra stipulations we need. For MAR data we need to look at potential drivers, and the different strengths (how strongly does this driver predict missingness). Similarly for MNAR we need to consider the strength and direction of missing data. 

The following is the Analysis function - it applies the causal method onto a given dataset and applies the MI. 
```{r updated_analysis}
Analyze_WithEmbeddedBalance <- function(condition, dat, fixed_objects = NULL) {
  cat("CLASS OF DAT:", class(dat), "| IS DF:", is.data.frame(dat), "| NAMES:", paste(names(dat), collapse=","), "\n")
  suppressPackageStartupMessages({
    library(survival)
    library(MASS)
    library(WeightIt)
    library(ebal)
    library(mice)
    library(dplyr)
  })
  
  if(is.null(condition) || is.null(dat) || is.null(fixed_objects))
    stop("ANALYZE ERROR: NULL inputs detected")
  
  pred_vars<- fixed_objects$pred.vars.processed
  cat_vars <- fixed_objects$cat.vars.processed
  cont_vars <- fixed_objects$cont.vars.processed
  causal_method <- condition$causal_method
  missing_method <- condition$missing_method
  in.mis.subj <- condition$input_mis_subj
  
  # Helper: run one analysis on one dataset 
  run_analysis <- function(dataset, label, force_complete_case = FALSE) {
    
    if(force_complete_case || missing_method == "complete_case" || in.mis.subj == 0) {
      # complete case analysis
      att.dfs <- ff.att.analysis(dataset, 
                                 outcome_type = "DFS",
                                 method = causal_method,
                                 pred_vars = pred_vars,
                                 include_balance = FALSE,
                                 include_diagnostics = FALSE)
      att.os  <- ff.att.analysis(dataset, 
                                 outcome_type = "OS",
                                 method = causal_method,
                                 pred_vars = pred_vars,
                                 include_balance = FALSE,
                                 include_diagnostics = FALSE)
      
    } else if(missing_method == "multiple_imputation") {
      # MI analysis we set iterations to 5 for now
      nSim <- 5
      
      mi_results <- enhanced_mi_analysis(
        dataset = dataset,
        pred_vars = pred_vars,
        cat_vars = cat_vars,
        cont_vars = cont_vars,
        causal_method = causal_method,
        in.mis.subj = in.mis.subj,
        nSim = nSim,
        sim_seed = as.integer(Sys.time()) + sample(1e6, 1)
      )
      
      if(is.null(mi_results)) {
        att.dfs <- ff.att.analysis(dataset, 
                                   outcome_type = "DFS",
                                   method = causal_method,
                                   pred_vars = pred_vars,
                                   include_balance = FALSE)
        att.os <- ff.att.analysis(dataset, 
                                  outcome_type = "OS",
                                  method = causal_method,
                                  pred_vars = pred_vars,
                                  include_balance = FALSE)
      } else {
        att.dfs <- list(
          hr = mi_results$att.pooled.dfs,
          log_hr = log(mi_results$att.pooled.dfs),
          se = mi_results$att.se.dfs,
          pvalue = mi_results$att.pvalue.dfs,
          ci_lower = exp(mi_results$ci_lower_dfs),
          ci_upper = exp(mi_results$ci_upper_dfs)
        )
        att.os <- list(
          hr = mi_results$att.pooled.os,
          log_hr = log(mi_results$att.pooled.os),
          se = mi_results$att.se.os,
          pvalue = mi_results$att.pvalue.os,
          ci_lower = exp(mi_results$ci_lower_os),
          ci_upper = exp(mi_results$ci_upper_os)
        )
      }
    } else {
      stop(paste("ANALYZE ERROR: Unknown missing_method:", missing_method))
    }
    
    # Validate
    for(res in list(att.dfs, att.os)) {
      for(field in c("hr", "se", "pvalue", "ci_lower", "ci_upper")) {
        if(is.null(res[[field]]) || !is.finite(res[[field]]))
          stop(paste("Invalid", field, "in", label, "result"))
      }
      if(res$ci_lower <= 0 || res$ci_upper <= 0)
        stop(paste("CI must be positive (HR scale) in", label))
      if(res$ci_lower >= res$ci_upper)
        stop(paste("CI lower >= upper in", label))
    }
    
    list(dfs = att.dfs, os = att.os)
  }
  
  # Oracle always uses complete case
  oracle <- run_analysis(dat$dataset.complete, "oracle", force_complete_case = TRUE)
  
  # Missing datasets use whatever missing_method specifies
  if(in.mis.subj > 0) {
    res_mcar <- run_analysis(dat$dataset.mcar, "mcar")
    res_mar <- run_analysis(dat$dataset.mar,  "mar")
    res_mnar <- run_analysis(dat$dataset.mnar, "mnar")
  } else {
    res_mcar <- oracle
    res_mar <- oracle
    res_mnar <- oracle
  }
  
  # Package into single results row 
  data.frame(
    # Oracle (complete data) estimates
    oracle.hr.dfsb= oracle$dfs$hr,
    oracle.hr.os = oracle$os$hr,
    
    # MCAR estimates
    mcar.hr.dfs = res_mcar$dfs$hr,
    mcar.hr.os = res_mcar$os$hr,
    mcar.se.dfs = res_mcar$dfs$se,
    mcar.se.os = res_mcar$os$se,
    mcar.pvalue.dfs = res_mcar$dfs$pvalue,
    mcar.pvalue.os = res_mcar$os$pvalue,
    mcar.ci.lower.dfs = res_mcar$dfs$ci_lower,
    mcar.ci.upper.dfs = res_mcar$dfs$ci_upper,
    mcar.ci.lower.os = res_mcar$os$ci_lower,
    mcar.ci.upper.os = res_mcar$os$ci_upper,
    
    # MAR estimates
    mar.hr.dfs = res_mar$dfs$hr,
    mar.hr.os = res_mar$os$hr,
    mar.se.dfs = res_mar$dfs$se,
    mar.se.os = res_mar$os$se,
    mar.pvalue.dfs = res_mar$dfs$pvalue,
    mar.pvalue.os = res_mar$os$pvalue,
    mar.ci.lower.dfs = res_mar$dfs$ci_lower,
    mar.ci.upper.dfs = res_mar$dfs$ci_upper,
    mar.ci.lower.os = res_mar$os$ci_lower,
    mar.ci.upper.os = res_mar$os$ci_upper,
    
    # MNAR estimates
    mnar.hr.dfs = res_mnar$dfs$hr,
    mnar.hr.os = res_mnar$os$hr,
    mnar.se.dfs = res_mnar$dfs$se,
    mnar.se.os = res_mnar$os$se,
    mnar.pvalue.dfs = res_mnar$dfs$pvalue,
    mnar.pvalue.os = res_mnar$os$pvalue,
    mnar.ci.lower.dfs = res_mnar$dfs$ci_lower,
    mnar.ci.upper.dfs = res_mnar$dfs$ci_upper,
    mnar.ci.lower.os = res_mnar$os$ci_lower,
    mnar.ci.upper.os = res_mnar$os$ci_upper,
    
    stringsAsFactors = FALSE
  )
}
```

The following is a 2nd version of the analysis function that is used with the 2n version of generate. For now it is not needed or used unless the first version somehow blows up. This version assumes Generate only makes the complete dataset and hence the missingness is done here instead. 
```{r analyze_v.2}
Analyze_WithEmbeddedBalance2 <- function(condition, dat, fixed_objects = NULL) {
  
  suppressPackageStartupMessages({
    library(survival); library(MASS)
    library(WeightIt); library(ebal)
    library(mice); library(dplyr)
  })
  
  # dat IS dataset.complete, passed directly from Generate 
  cat("CLASS OF DAT:", class(dat), 
      "| NROW:", nrow(dat), 
      "| NCOL:", ncol(dat), "\n")
  
  if(!is.data.frame(dat) || is.null(colnames(dat)) || nrow(dat) == 0)
    stop(paste("ANALYZE ERROR: dat is not a valid dataframe.",
               "Class:", class(dat), "Nrow:", nrow(dat)))
  
  pred_vars <- fixed_objects$pred.vars.processed
  cat_vars <- fixed_objects$cat.vars.processed
  cont_vars <- fixed_objects$cont.vars.processed
  causal_method <- condition$causal_method
  missing_method <- condition$missing_method
  in.mis.subj <- condition$input_mis_subj
  miss_vars <- condition$miss_vars
  if(is.list(miss_vars)) miss_vars <- unlist(miss_vars)
  
  # dat is always the complete dataset
  dataset.complete <- dat
  
  # Rebuild missing datasets inside Analyze
  # This avoids SimDesign list-unwrapping issues entirely
  if(in.mis.subj > 0) {
    
    dataset.mcar <- ff.miss.sub.mcar(
      dataset = dataset.complete,
      in.subj = in.mis.subj,
      in.pred = miss_vars
    )
    dataset.mar <- ff.miss.mar(
      dataset = dataset.complete,
      in.subj = in.mis.subj,
      in.pred = miss_vars,
      drivers = fixed_objects$mar_drivers[miss_vars]
    )
    dataset.mnar <- ff.miss.mnar(
      dataset = dataset.complete,
      in.subj = in.mis.subj,
      in.pred = miss_vars
    )
    
    # Validate missing datasets
    for(nm in c("dataset.mcar", "dataset.mar", "dataset.mnar")) {
      ds <- get(nm)
      if(is.null(colnames(ds)) || nrow(ds) == 0)
        stop(paste("ANALYZE ERROR:", nm, "invalid after missingness application"))
      cat(nm, "missingness applied — NAs in miss_vars:",
          sum(is.na(ds[, miss_vars])), "\n")
    }
    
  } else {
    dataset.mcar <- dataset.complete
    dataset.mar <- dataset.complete
    dataset.mnar <- dataset.complete
  }
  
  # Helper: run one analysis on one dataset
  run_analysis <- function(dataset, label) {
      cat("run_analysis called with label:", label, "\n")
    cat("dataset class:", class(dataset), "\n")
    cat("dataset nrow:", nrow(dataset), "\n")
    cat("dataset colnames:", paste(colnames(dataset), collapse = ", "), "\n")
    
    if(missing_method == "complete_case" || in.mis.subj == 0) {
      
      att.dfs <- ff.att.analysis(dataset, 
                                 outcome_type = "DFS",
                                 method = causal_method,
                                 pred_vars = pred_vars,
                                 include_balance = FALSE,
                                 include_diagnostics = FALSE)
      att.os  <- ff.att.analysis(dataset, 
                                 outcome_type = "OS",
                                 method = causal_method,
                                 pred_vars = pred_vars,
                                 include_balance = FALSE,
                                 include_diagnostics = FALSE)
      
    } else if(missing_method == "multiple_imputation") {
      
      nSim <- max(20, round(in.mis.subj * 100))
      
      mi_results <- enhanced_parallel_mi_analysis(
        dataset = dataset,
        pred_vars = pred_vars,
        cat_vars = cat_vars,
        cont_vars = cont_vars,
        causal_method = causal_method,
        in.mis.subj = in.mis.subj,
        nSim = nSim,
        sim_seed = as.integer(Sys.time()) + sample(1e6, 1)
      )
      
      if(is.null(mi_results)) {
        # No missingness found - fall back to complete case
        att.dfs <- ff.att.analysis(dataset, 
                                   outcome_type = "DFS",
                                   method = causal_method, 
                                   pred_vars = pred_vars,
                                   include_balance = FALSE)
        att.os <- ff.att.analysis(dataset, 
                                  outcome_type = "OS",
                                  method = causal_method, 
                                  pred_vars = pred_vars,
                                  include_balance = FALSE)
      } else {
        att.dfs <- list(
          hr = mi_results$att.pooled.dfs,
          log_hr = log(mi_results$att.pooled.dfs),
          se = mi_results$att.se.dfs,
          pvalue = mi_results$att.pvalue.dfs,
          ci_lower = exp(mi_results$ci_lower_dfs),
          ci_upper = exp(mi_results$ci_upper_dfs)
        )
        att.os <- list(
          hr = mi_results$att.pooled.os,
          log_hr = log(mi_results$att.pooled.os),
          se = mi_results$att.se.os,
          pvalue = mi_results$att.pvalue.os,
          ci_lower = exp(mi_results$ci_lower_os),
          ci_upper = exp(mi_results$ci_upper_os)
        )
      }
      
    } else {
      stop(paste("ANALYZE ERROR: Unknown missing_method:", missing_method))
    }
    
    # Validate
    for(res in list(att.dfs, att.os)) {
      for(field in c("hr", "se", "pvalue", "ci_lower", "ci_upper")) {
        if(is.null(res[[field]]) || !is.finite(res[[field]]))
          stop(paste("Invalid", field, "in", label, "result"))
      }
      if(res$ci_lower <= 0 || res$ci_upper <= 0)
        stop(paste("CI must be positive (HR scale) in", label))
      if(res$ci_lower >= res$ci_upper)
        stop(paste("CI lower >= upper in", label))
    }
    
    list(dfs = att.dfs, os = att.os)
  }

  # Oracle: run on complete data 
  oracle <- run_analysis(dataset.complete, "oracle")
  
  # Run on all three missing datasets 
  if(in.mis.subj > 0) {
    res_mcar <- run_analysis(dataset.mcar, "mcar")
    res_mar <- run_analysis(dataset.mar,  "mar")
    res_mnar <- run_analysis(dataset.mnar, "mnar")
  } else {
    res_mcar <- oracle
    res_mar <- oracle
    res_mnar <- oracle
  }
  
  # Package into single results row 
  data.frame(
    # Oracle (complete data) estimates
    oracle.hr.dfs = oracle$dfs$hr,
    oracle.hr.os = oracle$os$hr,
    
    # MCAR estimates
    mcar.hr.dfs = res_mcar$dfs$hr,
    mcar.hr.os = res_mcar$os$hr,
    mcar.se.dfs = res_mcar$dfs$se,
    mcar.se.os = res_mcar$os$se,
    mcar.pvalue.dfs = res_mcar$dfs$pvalue,
    mcar.pvalue.os = res_mcar$os$pvalue,
    mcar.ci.lower.dfs = res_mcar$dfs$ci_lower,
    mcar.ci.upper.dfs = res_mcar$dfs$ci_upper,
    mcar.ci.lower.os = res_mcar$os$ci_lower,
    mcar.ci.upper.os = res_mcar$os$ci_upper,
    
    # MAR estimates
    mar.hr.dfs = res_mar$dfs$hr,
    mar.hr.os = res_mar$os$hr,
    mar.se.dfs = res_mar$dfs$se,
    mar.se.os = res_mar$os$se,
    mar.pvalue.dfs = res_mar$dfs$pvalue,
    mar.pvalue.os = res_mar$os$pvalue,
    mar.ci.lower.dfs = res_mar$dfs$ci_lower,
    mar.ci.upper.dfs = res_mar$dfs$ci_upper,
    mar.ci.lower.os = res_mar$os$ci_lower,
    mar.ci.upper.os = res_mar$os$ci_upper,
    
    # MNAR estimates
    mnar.hr.dfs = res_mnar$dfs$hr,
    mnar.hr.os = res_mnar$os$hr,
    mnar.se.dfs = res_mnar$dfs$se,
    mnar.se.os = res_mnar$os$se,
    mnar.pvalue.dfs = res_mnar$dfs$pvalue,
    mnar.pvalue.os = res_mnar$os$pvalue,
    mnar.ci.lower.dfs = res_mnar$dfs$ci_lower,
    mnar.ci.upper.dfs = res_mnar$dfs$ci_upper,
    mnar.ci.lower.os = res_mnar$os$ci_lower,
    mnar.ci.upper.os = res_mnar$os$ci_upper,
    
    stringsAsFactors = FALSE
  )
}
```

The final part of the DesignSim - the summarise function. This is pretty simple - it just creates the summary statistics we want for each simulation condition. This usually means just taking the mean/sd over the replicants. 
```{r, summarise_results}
Summarise <- function(condition, results, fixed_objects = NULL) {
  
  true_log_hr_dfs <- log(condition$input_hr_dfs)
  true_log_hr_os <- log(condition$input_hr_os)
  
  # Helper to compute all metrics for one mechanism
  compute_metrics <- function(hr_dfs, 
                              hr_os, 
                              ci_lower_dfs, 
                              ci_upper_dfs,
                              ci_lower_os,  
                              ci_upper_os,
                              pvalue_dfs,   
                              pvalue_os,
                              prefix) {
    
    bias_dfs <- log(hr_dfs) - true_log_hr_dfs
    bias_os <- log(hr_os) - true_log_hr_os
    
    setNames(data.frame(
      # HR estimates 
      mean(hr_dfs, na.rm = TRUE),
      median(hr_dfs, na.rm = TRUE),
      mean(hr_os, na.rm = TRUE),
      median(hr_os, na.rm = TRUE),
      
      # DFS metrics 
      mean(bias_dfs, na.rm = TRUE),
      sqrt(mean(bias_dfs^2, na.rm = TRUE)),
      mean(ci_lower_dfs <= condition$input_hr_dfs &
           ci_upper_dfs >= condition$input_hr_dfs, na.rm = TRUE),
      mean(pvalue_dfs < 0.05, na.rm = TRUE),
      
      # OS metrics 
      mean(bias_os, na.rm = TRUE),
      sqrt(mean(bias_os^2, na.rm = TRUE)),
      mean(ci_lower_os <= condition$input_hr_os &
           ci_upper_os >= condition$input_hr_os, na.rm = TRUE),
      mean(pvalue_os < 0.05,  na.rm = TRUE)
      
    ), paste0(prefix, c(
      ".mean.hr.dfs",   
      ".median.hr.dfs",
      ".mean.hr.os",    
      ".median.hr.os",
      ".bias.dfs",      
      ".rmse.dfs",
      ".coverage.dfs",  
      ".power.dfs",
      ".bias.os",       
      ".rmse.os",
      ".coverage.os",   
      ".power.os"
    )))
  }
  
  cbind(
    data.frame(
      true.hr.dfs = condition$input_hr_dfs,
      true.hr.os = condition$input_hr_os,
      n_replications = nrow(results)
    ),
    compute_metrics(
      results$oracle.hr.dfs, results$oracle.hr.os,
      results$mcar.ci.lower.dfs, results$mcar.ci.upper.dfs,
      results$mcar.ci.lower.os, results$mcar.ci.upper.os,
      results$mcar.pvalue.dfs, results$mcar.pvalue.os,
      prefix = "oracle"
    ),
    compute_metrics(
      results$mcar.hr.dfs, results$mcar.hr.os,
      results$mcar.ci.lower.dfs, results$mcar.ci.upper.dfs,
      results$mcar.ci.lower.os, results$mcar.ci.upper.os,
      results$mcar.pvalue.dfs, results$mcar.pvalue.os,
      prefix = "mcar"
    ),
    compute_metrics(
      results$mar.hr.dfs, results$mar.hr.os,
      results$mar.ci.lower.dfs, results$mar.ci.upper.dfs,
      results$mar.ci.lower.os, results$mar.ci.upper.os,
      results$mar.pvalue.dfs, results$mar.pvalue.os,
      prefix = "mar"
    ),
    compute_metrics(
      results$mnar.hr.dfs, results$mnar.hr.os,
      results$mnar.ci.lower.dfs, results$mnar.ci.upper.dfs,
      results$mnar.ci.lower.os, results$mnar.ci.upper.os,
      results$mnar.pvalue.dfs, results$mnar.pvalue.os,
      prefix = "mnar"
    )
  )
}
```

With this we are done with the simulation design code. All that is left is running the simulation on specificed conditions. This is given by the `Design` function. The tricky part is deciding what we want for our Design. 
```{r}
empirical_dist <- ff.mean.sigma(ref.data.processed)
baseline_models <- estimate_baseline_models(data)
treatment_models <- estimate_treatment_models(data)

# MAR drivers defined once here, passed to all replicates
mar_drivers <- list(
  "ENE_bin" = c("Total_LN_positive_", "pN_7th_bin"),
  "Total_LN_positive_" = c("Max_tumor_size", "pT_7th_bin"),
  "Age_Dx_bin" = c("pT_7th_bin", "PORT_bin"),
  "pN_7th_bin" = c("Total_LN_positive_", "Max_tumor_size"),
  "PORT_bin" = c("pN_7th_bin", "ENE_bin"),
  "Max_tumor_size" = c("pT_7th_bin", "pN_7th_bin"),
  "pT_7th_bin" = c("Max_tumor_size", "Age_Dx_bin")
)

# Fixed objects 
fixed_objects <- list(
  empirical_dist  = empirical_dist,
  baseline_models = baseline_models,
  treatment_models = treatment_models,
  pred.vars.processed = pred.vars.processed,
  cat.vars.processed  = cat.vars.processed,
  cont.vars.processed = cont.vars.processed,
  original_data = ref.data.processed,
  mar_drivers = mar_drivers
)

for(obj_name in names(fixed_objects)) {
  if(is.null(fixed_objects[[obj_name]])) {
    stop(paste("FIXED_OBJECTS ERROR:", obj_name, "is NULL"))
  }
}
cat("Fixed objects validated successfully\n")

############################## Simulation design ##############################
# missing_mechanism removed from condition grid — 
# Generate now produces all three missing datasets per replicate
# Analyze runs all three mechanisms in one pass
# This reduces conditions from 243 to 54 and ensures 
# MCAR/MAR/MNAR comparisons are within-replicate (same base dataset)

# An example of a final design might look like this:
Design <- createDesign(
  input_ss       = c(500, 1000, 2000),          # sample sizes
  input_hr_dfs   = c(1.0, 1.5, 2.5),            # DFS hazard ratios
  input_hr_os    = c(1.0, 1.3, 1.7),            # OS hazard ratios
  input_mis_subj = c(0, 0.25, 0.50),            # missingness proportions
  causal_method  = c("match", "ebal", "iptw"),     # causal methods
  missing_method = c("complete_case",
                     "multiple_imputation"),    # missing data analysis
  miss_vars      = list(
    c("ENE_bin", "Total_LN_positive_")),          # strongest confounders
  treatment_type = "engraftment"
)

if(is.null(Design) || nrow(Design) == 0) {
  stop("DESIGN ERROR: Failed to create simulation design matrix")
}
```

```{r}
# After preprocessing and model fitting — run this once
saveRDS(fixed_objects, "env/fixed_objects.rds")
saveRDS(empirical_dist, "env/empirical_dist.rds")
saveRDS(baseline_models, "env/baseline_models.rds")
saveRDS(treatment_models, "env/treatment_models.rds")
saveRDS(original_data, "env/original_data.rds")
saveRDS(Design, "env/Design.rds")
saveRDS(pred.vars.processed, "env/pred_vars.rds")
saveRDS(cat.vars.processed, "env/cat_vars.rds")
saveRDS(cont.vars.processed, "env/cont_vars.rds")

cat("All objects saved\n")
```


Before we run the entire simulation we can do a test run first. 

```{r}
# Test Generate on one condition
test_condition <- Design[1, ]

test_data <- Generate(test_condition, fixed_objects)

# Check all four datasets came back
names(test_data)  # should show complete, mcar, mar, mnar

# Check dimensions and missingness
nrow(test_data$dataset.complete)
colSums(is.na(test_data$dataset.mcar))
colSums(is.na(test_data$dataset.mar))
colSums(is.na(test_data$dataset.mnar))
```

######

Results for the poster. This section will the simulation study done for the poster (smaller subset of results). 

```{r}
Design_poster <- createDesign(
  input_ss       = 500,               
  input_hr_dfs   = c(1.0, 1.5, 2.0),       
  input_hr_os    = c(1.0, 1.5, 2.0),
  input_mis_subj = c(0, 0.25, 0.50),
  causal_method  = c("iptw", "ebal", "match"),
  missing_method = c("complete_case", "multiple_imputation"),
  miss_vars      = "ENE_bin, Total_LN_positive_",
  treatment_type = "engraftment"
)

cat("Total conditions:", nrow(Design_poster), "\n")
design_poster_cc <- Design_poster[
  Design_poster$missing_method == "complete_case", ]
design_poster_mi <- Design_poster[
  Design_poster$missing_method == "multiple_imputation" &
  Design_poster$input_mis_subj  > 0, ]

cat("CC conditions: ", nrow(design_poster_cc), "\n")  # 36
cat("MI conditions: ", nrow(design_poster_mi), "\n")  # 24
```

Once we have a design, the `runSimulation` function will do everything for us. Recommendtion: for larger cases it might be better to break up the simulation. For example, between CC and MI parts to speed up runtime and avoid any issues of crashing overnight.
```{r design_simulation_CC}
SimClean()
results_poster_cc <- runSimulation(
  design        = design_poster_cc,
  replications  = 200,
  generate      = Generate,
  analyse       = Analyze_WithEmbeddedBalance,
  summarise     = Summarise,
  fixed_objects = fixed_objects,
  parallel      = TRUE,
  ncores        = parallel::detectCores() - 1,
  save          = TRUE,
  save_results  = TRUE,
  filename      = "outputs/poster_cc",
  verbose       = TRUE
)

filename_final_poster_cc <- paste0("outputs/poster_cc.csv")
write.csv(results_poster_cc, filename_final_poster_cc)
```

```{r design_simulation_MI}
SimClean()
results_poster_mi <- runSimulation(
  design        = design_poster_mi,
  replications  = 200,             
  generate      = Generate,
  analyse       = Analyze_WithEmbeddedBalance,
  summarise     = Summarise,
  fixed_objects = fixed_objects,
  parallel      = TRUE,
  ncores        = parallel::detectCores() - 1,
  save          = TRUE,
  save_results  = TRUE,
  filename      = "outputs/poster_mi_long",
  verbose       = TRUE
)

filename_final_poster_mi <- paste0("outputs/poster_mi_long.csv")
write.csv(results_poster_mi_long, filename_final_poster_mi)
```

```{r design_simulation_FULL}
# Combine all
results_poster_full <- rbind(
  results_poster_cc,
  results_poster_mi
)
saveRDS(results_poster_full, "outputs/poster_full_final.rds")
filename_final_poster <- paste0("outputs/poster_data.csv")
write.csv(results_poster_full, filename_final_poster)
```


####
```{r}
#Save work to excel
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
filename <- paste0("outputs/pilot_oscc_simulation_", timestamp, ".xlsx")
write_xlsx(pilot_results, filename)
```


What the full simulation might look like. 
```{r}
#SimClean()
full_results <- runSimulation(
  design        = Design,
  replications  = 100,
  generate      = Generate,
  analyse       = Analyze_WithEmbeddedBalance,
  summarise     = Summarise,
  fixed_objects = fixed_objects,
  parallel      = TRUE,
  ncores        = parallel::detectCores() - 1,
  save          = TRUE,
  save_results  = TRUE,
  filename      = "full_simulation_results",
  verbose       = TRUE
)
```


```{r}
#Save work to excel
# simulation broke
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
filename_final <- paste0("outputs/full_oscc_simulation_", timestamp, ".xlsx")
write_xlsx(full_simulation_results, filename_final)
```



